---
title: "Microplastic - 03: Alpha Diversity"
author: "Jan Waelchli"
geometry: margin=2cm
output:
  pdf_document:
    toc: yes
    toc_depth: 3
---

\pagebreak

```{r setup, include=FALSE, echo=F, warning=F, message=F,}

##clear the object from memory
rm(list=ls())

#knitr settings
knitr::opts_chunk$set(echo=TRUE, fig.align="center")
options(tinytex.verbose = TRUE)

#set seed
set.seed(100)

#paths
paths <- list(functions="functions/")

## set source
library(rstudioapi)
setwd(dirname(getActiveDocumentContext()$path))

## load functions
source(paste0(paths$functions,"functions.R"))
functions(path=paths$functions)

## installs (if necessary) and loads libraries
libraries()
library(tidyverse)
library(lme4)

```

```{r import, echo=F, message=F, warning=F}

# Import

# set wd to RDS files
setwd("interim/01_Import_Normalization")

#Import RDS files
rare_threshold <- readRDS("rare_threshold.RDS")

DESIGN <- readRDS("DESIGN.RDS")
DAT <- readRDS("DAT.RDS")
DAT_rare <- readRDS("DAT_rare.RDS")

DESIGN_sand <- readRDS("DESIGN_sand.RDS")
DAT_sand <- readRDS("DAT_sand.RDS")
DAT_sand_rare <- readRDS("DAT_sand_rare.RDS")

DESIGN_root_sand <- readRDS("DESIGN_root_sand.RDS")
DAT_root_sand <- readRDS("DAT_root_sand.RDS")
DAT_root_sand_rare <- readRDS("DAT_root_sand_rare.RDS")

DESIGN_root_soil <- readRDS("DESIGN_root_soil.RDS")
DAT_root_soil <- readRDS("DAT_root_soil.RDS")
DAT_root_soil_rare <- readRDS("DAT_root_soil_rare.RDS")

DESIGN_soil <- readRDS("DESIGN_soil.RDS")
DAT_soil <- readRDS("DAT_soil.RDS")
DAT_soil_rare <- readRDS("DAT_soil_rare.RDS")

level_cols_Plastic <- readRDS("level_cols_Plastic.RDS")
level_cols_Substrate <- readRDS("level_cols_Substrate.RDS")

```

# Alpha diversity

We answer the following question for the alpha diversity in each substrate:

-   **Q1: Are there differences in alpha diversity between the different treatments?**
-   **Q2: Are there differences in alpha diversity between the different concentrations?**
<!-- -   **Q3: Are there differences in alpha diversity between the samples with plant and samples without plants?** -->

\vspace{10mm}

## Rarefaction

### Figure 3 \| Rarefaction

Before analyzing the alpha diversity, we conduct the rarefaction plot to make sure that no diversity is lost due to a too low rarefaction threshold.

\vspace{5mm}

```{r rarefaction, eval=T, echo=F, message=F, warning=F, fig.height=5, fig.width=6}

#rarefaction all substrate

pdf(NULL) #prevents drawing of standard rarefaction plot (we will create our own with colors)
rarefaction <- rarecurve(t(DAT), step=200, 7000)
invisible(dev.off())

#png("rarefaction.png",4000, 3200, res = 600)

#limit for plotting
bSmax <- sapply(rarefaction, max)

### BACTERIA ###
#plot rarefaction
par(mar=c(5.1, 4.1, 2.1, 2.1), xpd=F)#add extra space on the rigth side
plot(c(1, 50000), c(1, max(bSmax)), xlab="Sequencing depth", main="Bacteria: Rarefaction",
     ylab="number of ASVs", type="n", las=1, ylim=c(0,600))
for (i in seq_along(rarefaction)) {
  N <- attr(rarefaction[[i]], "Subsample")
  lines(N, rarefaction[[i]], col=DESIGN$cols_Substrate[i], lwd=0.5)
}
#add rarefaction threshold
abline(v=rare_threshold, lty = "longdash", lwd=2, col="red")

legend("bottomright", title.adj = 0.2, title = "Substrate", inset=c(0, 0),
       legend=names(level_cols_Substrate),
       pch=16, bty = "n", col=level_cols_Substrate)

```


```{r rarefaction subplots, eval=T, echo=F, message=F, warning=F, fig.height=5, fig.width=6}

#function to plot subsets
rarefaction_plot <- function(substrate){
  #get subseted data & design
  data_sub <- get(paste0("DAT_", tolower(substrate)))
  design_sub <- get(paste0("DESIGN_", tolower(substrate)))
  #subset colorstring
  level_cols_sub <- level_cols_Substrate[grepl(paste0("^", substrate), names(level_cols_Substrate), ignore.case = T)]
  #rarefy
  pdf(NULL) #prevents drawing of standard rarefaction plot (we will create our own with colors)
  rarefaction <- rarecurve(t(data_sub), step=200, 7000)
  invisible(dev.off())
  #limit for plotting
  bSmax <- sapply(rarefaction, max)
  #plot rarefaction
  par(mar=c(5.1, 4.1, 2.1, 2.1), xpd=F)#add extra space on the rigth side
  plot(c(1, 50000), c(1, max(bSmax)), xlab="Sequencing depth", main="Rarefaction",
       ylab="number of ASVs", type="n", las=1, ylim=c(0,600))
  for (i in seq_along(rarefaction)) {
    N <- attr(rarefaction[[i]], "Subsample")
    lines(N, rarefaction[[i]], col=design_sub$cols_Substrate[i], lwd=0.5)
  }
  #add rarefaction threshold
  abline(v=rare_threshold, lty = "longdash", lwd=2, col="red")
  legend("bottomright", title.adj = 0.2, title = "Substrate", inset=c(0, 0),
         legend=names(level_cols_sub),
         pch=16, bty = "n", col=level_cols_sub)
  
}


#make plots
rarefaction_plot("Soil")
rarefaction_plot("Root_soil")
rarefaction_plot("Root_sand")
rarefaction_plot("Sand")

```

\vspace{5mm}

**Conclusion:** We chose a rarefaction threshold that do cover the majority of diversity in each sample.

\vspace{10mm}
\pagebreak

## Alpha Diversity

Method: We rarefy the dataset by the sequencing depth of `r rare_threshold` and calculate the Shannon diversity for each sample. This is repeated 100 times. Then, the mean value from the 100 iterations is taken for statistical analysis between the different samples.

```{r a-div shannon, warning=F, echo=F, message=F, eval=T}

####
#easier to write, from here on we only use the rarefied datasets
DAT_sand <- DAT_sand_rare
DAT_root_sand <- DAT_root_sand_rare
DAT_root_soil <- DAT_root_soil_rare
DAT_soil <- DAT_soil_rare
####

repetitions <- 1

#sand

  #rarefy data a 100 time and calculate each time the alpha diversity (shannon)
  #differs each time because rarefaction is a random process
  diversity_DAT_sand_mat <- c()
  for (i in 1:repetitions){
    #generate rarefied community
    temp_mat <- t(rrarefy(t(DAT_sand), rare_threshold))
    #calculate shannon
    shannon <- vegan::diversity(temp_mat, index="shannon", MARGIN=2)
    d <- exp(shannon)
    #combine
    diversity_DAT_sand_mat <- cbind(diversity_DAT_sand_mat, d)
  }

#calculate means of subsamples
asv_diversity <- rowMeans(diversity_DAT_sand_mat)
#put them all in one dataframe
alpha_diversity_DAT_sand_summary <- cbind(asv_diversity, DESIGN_sand)
rm(diversity_DAT_sand_mat, alpha_diversity)


#root_sand

  #rarefy data a 100 time and calculate each time the alpha diversity (shannon)
  #differs each time because rarefaction is a random process
  diversity_DAT_root_sand_mat <- c()
  for (i in 1:repetitions){
    #generate rarefied community
    temp_mat <- t(rrarefy(t(DAT_root_sand), rare_threshold))
    #calculate shannon
    shannon <- vegan::diversity(temp_mat, index="shannon", MARGIN=2)
    d <- exp(shannon)
    #combine
    diversity_DAT_root_sand_mat <- cbind(diversity_DAT_root_sand_mat, d)
  }

#calculate means of subsamples
asv_diversity <- rowMeans(diversity_DAT_root_sand_mat)
#put them all in one dataframe
alpha_diversity_DAT_root_sand_summary <- cbind(asv_diversity, DESIGN_root_sand)
rm(diversity_DAT_root_sand_mat, alpha_diversity)


#root_soil

  #rarefy data a 100 time and calculate each time the alpha diversity (shannon)
  #differs each time because rarefaction is a random process
  diversity_DAT_root_soil_mat <- c()
  for (i in 1:repetitions){
    #generate rarefied community
    temp_mat <- t(rrarefy(t(DAT_root_soil), rare_threshold))
    #calculate shannon
    shannon <- vegan::diversity(temp_mat, index="shannon", MARGIN=2)
    d <- exp(shannon)
    #combine
    diversity_DAT_root_soil_mat <- cbind(diversity_DAT_root_soil_mat, d)
  }

#calculate means of subsamples
asv_diversity <- rowMeans(diversity_DAT_root_soil_mat)
#put them all in one dataframe
alpha_diversity_DAT_root_soil_summary <- cbind(asv_diversity, DESIGN_root_soil)
rm(diversity_DAT_root_soil_mat, alpha_diversity)


#soil

  #rarefy data a 100 time and calculate each time the alpha diversity (shannon)
  #differs each time because rarefaction is a random process
  diversity_DAT_soil_mat <- c()
  for (i in 1:repetitions){
    #generate rarefied community
    temp_mat <- t(rrarefy(t(DAT_soil), rare_threshold))
    #calculate shannon
    shannon <- vegan::diversity(temp_mat, index="shannon", MARGIN=2)
    d <- exp(shannon)
    #combine
    diversity_DAT_soil_mat <- cbind(diversity_DAT_soil_mat, d)
  }

#calculate means of subsamples
asv_diversity <- rowMeans(diversity_DAT_soil_mat)
#put them all in one dataframe
alpha_diversity_DAT_soil_summary <- cbind(asv_diversity, DESIGN_soil)
rm(diversity_DAT_soil_mat, alpha_diversity)

```

\vspace{10mm}

## Overview

We get an overview for each substrate by investigating the effect on alpha diversity by the factors of plastic, concentration and the interaction between them. We model the alpha diversity against these factors in an aov-model and perform a F-Test.

```{r a-div all bacteria tp * hc in root, warning=F, echo=F, message=F, eval=T}

#sand
alpha_sand_aov <- aov(asv_diversity ~ Plastic * Concentration + Plant, data=alpha_diversity_DAT_sand_summary)
pander(summary(alpha_sand_aov), caption = "Sand: F test")

#root_sand
alpha_root_sand_aov <- aov(asv_diversity ~ Plastic * Concentration, data=alpha_diversity_DAT_root_sand_summary)
pander(summary(alpha_root_sand_aov), caption = "Root_sand: F test")

#root_soil
alpha_root_soil_aov <- aov(asv_diversity ~ Plastic * Concentration, data=alpha_diversity_DAT_root_soil_summary)
pander(summary(alpha_root_soil_aov), caption = "Root_soil: F test")

#Soil
alpha_soil_aov <- aov(asv_diversity ~ Plastic * Concentration + Plant, data=alpha_diversity_DAT_soil_summary)
pander(summary(alpha_soil_aov), caption = "Soil: F test")

```

**Conclusion:** We find an effect on the alpha diversity between the different plastics and for the plastic concentration in sand.

\pagebreak

## Plastic effect

**Q1: Are there differences in alpha diversity between the different treatments (samples with plant)?**\
For this analysis we only conduct samples with plants. Plants were either treated with no plastic (control), microplastic, macroplastic or starch. We saw above that the treatment is shaping the alpha diversity. We investigate now in which direction is the alpha diversity shifted by the single treatments by looking at Shannon diversity, richness and evenness.

### Shannon Diversity

```{r a-div Plastic shanon, warning=F, echo=F, message=F, eval=T}

pairwise_test_Plastic <- function(aov_data, summary_data){
  alpha_aov_tukey <- emmeans(aov_data, c("Plastic"))  # library(emmeans)
  alpha_aov_tukey_letters <- cld(alpha_aov_tukey, Letter="abcdefghi", alpha=0.05)
  # means
  alpha_means <- aggregate(summary_data$asv_diversity,list(summary_data$Plastic), mean)
  # table
  alpha_aov_tukey_letters <- alpha_aov_tukey_letters[order(with(alpha_aov_tukey_letters, Plastic)),]
  alpha_means <- alpha_means[order(with(alpha_means, Group.1, Group.2)),]
  alpha_table <- data.frame(Plastic=alpha_means[,1], alphaIndex=alpha_means[,2], letters=alpha_aov_tukey_letters[,7])
  return(alpha_table)
}

#sand
df <- alpha_diversity_DAT_sand_summary[alpha_diversity_DAT_sand_summary$Plant == T,] #subset for only plant samples
alpha_sand_aov <- aov(asv_diversity ~ Plastic, data=df)
alpha_table_Plastic_sand <- pairwise_test_Plastic(alpha_sand_aov, df)
pander(alpha_table_Plastic_sand, caption = "Plastic effect in sand")

#root_sand
df <- alpha_diversity_DAT_root_sand_summary[alpha_diversity_DAT_root_sand_summary$Plant == T,]
alpha_root_sand_aov <- aov(asv_diversity ~ Plastic, data=df)
alpha_table_Plastic_root_sand <- pairwise_test_Plastic(alpha_root_sand_aov, df)
pander(alpha_table_Plastic_root_sand, caption = "Plastic effect in root_sand")

#root_soil
df <- alpha_diversity_DAT_root_soil_summary[alpha_diversity_DAT_root_soil_summary$Plant == T,]
alpha_root_soil_aov <- aov(asv_diversity ~ Plastic, data=df)
alpha_table_Plastic_root_soil <- pairwise_test_Plastic(alpha_root_soil_aov, df)
pander(alpha_table_Plastic_root_soil, caption = "Plastic effect in root_soil")

#soil
df <- alpha_diversity_DAT_soil_summary[alpha_diversity_DAT_soil_summary$Plant == T,]
alpha_soil_aov <- aov(asv_diversity ~ Plastic, data=df)
alpha_table_Plastic_soil <- pairwise_test_Plastic(alpha_soil_aov, df)
pander(alpha_table_Plastic_soil, caption = "Plastic effect in soil")

```

### Figure 4.1 \| Plastic effect on alpha diversity
\vspace{5mm}

```{r plot a-div Plastic shanon, echo=F, message=F, warning=F, eval=T, , fig.height=5, fig.width=8}

#combine data.frames
alpha_diversity_DAT_summary <- rbind(alpha_diversity_DAT_sand_summary, alpha_diversity_DAT_root_sand_summary,
                                     alpha_diversity_DAT_root_soil_summary, alpha_diversity_DAT_soil_summary)
df <- alpha_diversity_DAT_summary[alpha_diversity_DAT_summary$Plant == T,]

#data frame for letters
alpha_table_Plastic <- rbind(alpha_table_Plastic_sand, alpha_table_Plastic_root_sand, alpha_table_Plastic_root_soil, alpha_table_Plastic_soil)
alpha_table_Plastic$Substrate <- rep(c("Sand", "Root_sand", "Root_soil", "Soil"), each=4)
alpha_table_Plastic$Substrate <- factor(alpha_table_Plastic$Substrate, levels=c("Sand", "Root_sand", "Root_soil", "Soil"))
alpha_table_Plastic$letters <- trimws(alpha_table_Plastic$letters) #remove whitespaces
alpha_table_Plastic$x <- rep(1:4, 4)
alpha_table_Plastic$y <- c(
tapply(alpha_diversity_DAT_sand_summary$asv_diversity, alpha_diversity_DAT_sand_summary$Plastic, max)+15,
tapply(alpha_diversity_DAT_root_sand_summary$asv_diversity, alpha_diversity_DAT_root_sand_summary$Plastic, max)+15,
tapply(alpha_diversity_DAT_root_soil_summary$asv_diversity, alpha_diversity_DAT_root_soil_summary$Plastic, max)+15,
tapply(alpha_diversity_DAT_soil_summary$asv_diversity, alpha_diversity_DAT_soil_summary$Plastic, max)+15)

#concentration
df$Concentration <- factor(paste0(df$Concentration, "%"), levels=c("0%", "3%", "10%"))

#plot
 
plot <- ggplot(df, aes(y=asv_diversity, x=Plastic, fill=Plastic)) +
            geom_boxplot(position=position_dodge2(width=0.75, preserve="single"),outlier.colour = NA) +
            geom_jitter(size=1, position=position_jitterdodge(jitter.width=0.3,dodge.width = 0.75), alpha=0.7, aes(col=Concentration))+
            theme_bw() +
            theme(legend.position="none") +
            ylab("Shanon Diversity")+
            xlab("")+
            scale_fill_manual(values=level_cols_Plastic) +
            scale_color_manual(values=c("black","dimgrey","grey")) +
            facet_grid(~ Substrate)+
            theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size=7), strip.text = element_text(size = 6)) +
            labs(title = "Bacteria: Alpha diversity", subtitle = "alpha diversity ~ Plastic + Plant")+
            geom_text(data = alpha_table_Plastic, aes(x=x, y=y, label=letters),size=4, col="dimgrey")



### PLOT

#combine the two plots (library cowplot)
fig4 <- plot_grid(plot + theme(legend.position="bottom"))

#print
#png("figures/fig4.png",6400, 4000, res = 600)
fig4
#dev.off()

```

\vspace{5mm}
**Conclusion:** In sand and, the alpha diversity is in overall reduced by treating with plastics. In roots from sand and soil, the opposite effect can be observed.

\pagebreak

### Richness

```{r a-div Plastic richness, warning=F, echo=F, message=F, eval=T}

#get richness
alpha_diversity_DAT_sand_summary$richness <- apply(DAT_sand_rare, 2, function(x) sum(x > 0))
alpha_diversity_DAT_root_sand_summary$richness <- apply(DAT_root_sand_rare, 2, function(x) sum(x > 0))
alpha_diversity_DAT_root_soil_summary$richness <- apply(DAT_root_soil_rare, 2, function(x) sum(x > 0))
alpha_diversity_DAT_soil_summary$richness <- apply(DAT_soil_rare, 2, function(x) sum(x > 0))

#sand
df <- alpha_diversity_DAT_sand_summary[alpha_diversity_DAT_sand_summary$Plant == T,] #subset for only plant samples
alpha_sand_aov <- aov(richness ~ Plastic, data=df)
alpha_table_Plastic_sand <- pairwise_test_Plastic(alpha_sand_aov, df)
pander(alpha_table_Plastic_sand, caption = "Plastic effect in sand")

#root_sand
df <- alpha_diversity_DAT_root_sand_summary[alpha_diversity_DAT_root_sand_summary$Plant == T,]
alpha_root_sand_aov <- aov(richness ~ Plastic, data=df)
alpha_table_Plastic_root_sand <- pairwise_test_Plastic(alpha_root_sand_aov, df)
pander(alpha_table_Plastic_root_sand, caption = "Plastic effect in root_sand")

#root_soil
df <- alpha_diversity_DAT_root_soil_summary[alpha_diversity_DAT_root_soil_summary$Plant == T,]
alpha_root_soil_aov <- aov(richness ~ Plastic, data=df)
alpha_table_Plastic_root_soil <- pairwise_test_Plastic(alpha_root_soil_aov, df)
pander(alpha_table_Plastic_root_soil, caption = "Plastic effect in root_soil")

#soil
df <- alpha_diversity_DAT_soil_summary[alpha_diversity_DAT_soil_summary$Plant == T,]
alpha_soil_aov <- aov(richness ~ Plastic, data=df)
alpha_table_Plastic_soil <- pairwise_test_Plastic(alpha_soil_aov, df)
pander(alpha_table_Plastic_soil, caption = "Plastic effect in soil")

```

```{r plot a-div Plastic richness, echo=F, message=F, warning=F, eval=T, , fig.height=5, fig.width=8}

#combine data.frames
alpha_diversity_DAT_summary <- rbind(alpha_diversity_DAT_sand_summary, alpha_diversity_DAT_root_sand_summary,
                                     alpha_diversity_DAT_root_soil_summary, alpha_diversity_DAT_soil_summary)
df <- alpha_diversity_DAT_summary[alpha_diversity_DAT_summary$Plant == T,]

#data frame for letters
alpha_table_Plastic <- rbind(alpha_table_Plastic_sand, alpha_table_Plastic_root_sand, alpha_table_Plastic_root_soil, alpha_table_Plastic_soil)
alpha_table_Plastic$Substrate <- rep(c("Sand", "Root_sand", "Root_soil", "Soil"), each=4)
alpha_table_Plastic$Substrate <- factor(alpha_table_Plastic$Substrate, levels=c("Sand", "Root_sand", "Root_soil", "Soil"))
alpha_table_Plastic$letters <- trimws(alpha_table_Plastic$letters) #remove whitespaces
alpha_table_Plastic$x <- rep(1:4, 4)
alpha_table_Plastic$y <- c(
tapply(alpha_diversity_DAT_sand_summary$richness, alpha_diversity_DAT_sand_summary$Plastic, max)+15,
tapply(alpha_diversity_DAT_root_sand_summary$richness, alpha_diversity_DAT_root_sand_summary$Plastic, max)+15,
tapply(alpha_diversity_DAT_root_soil_summary$richness, alpha_diversity_DAT_root_soil_summary$Plastic, max)+15,
tapply(alpha_diversity_DAT_soil_summary$richness, alpha_diversity_DAT_soil_summary$Plastic, max)+15)

#concentration
df$Concentration <- factor(paste0(df$Concentration, "%"), levels=c("0%", "3%", "10%"))

#plot
 
plot <- ggplot(df, aes(y=richness, x=Plastic, fill=Plastic)) +
            geom_boxplot(position=position_dodge2(width=0.75, preserve="single"),outlier.colour = NA) +
            geom_jitter(size=1, position=position_jitterdodge(jitter.width=0.3,dodge.width = 0.75), alpha=0.7, aes(col=Concentration))+
            theme_bw() +
            theme(legend.position="none") +
            ylab("Richness")+
            xlab("")+
            scale_fill_manual(values=level_cols_Plastic) +
            scale_color_manual(values=c("black","dimgrey","grey")) +
            facet_grid(~ Substrate)+
            theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size=7), strip.text = element_text(size = 6)) +
            labs(title = "Bacteria: Alpha diversity", subtitle = "alpha diversity ~ Plastic + Plant")+
            geom_text(data = alpha_table_Plastic, aes(x=x, y=y, label=letters),size=4, col="dimgrey")



### PLOT

#combine the two plots (library cowplot)
fig4 <- plot_grid(plot + theme(legend.position="bottom"))

#print
#png("figures/fig4.2.png",6400, 4000, res = 600)
fig4
#dev.off()

```


\pagebreak

### Pieulou's evenness

```{r a-div Plastic evenness, warning=F, echo=F, message=F, eval=T}

library(chemodiv)

#get evenness
alpha_diversity_DAT_sand_summary$evenness <- unlist(chemodiv::calcDiv(t(DAT_sand_rare), type = "PielouEven"))
alpha_diversity_DAT_root_sand_summary$evenness <- unlist(chemodiv::calcDiv(t(DAT_root_sand_rare), type = "PielouEven"))
alpha_diversity_DAT_root_soil_summary$evenness <- unlist(chemodiv::calcDiv(t(DAT_root_soil_rare), type = "PielouEven"))
alpha_diversity_DAT_soil_summary$evenness <- unlist(chemodiv::calcDiv(t(DAT_soil_rare), type = "PielouEven"))

#sand
df <- alpha_diversity_DAT_sand_summary[alpha_diversity_DAT_sand_summary$Plant == T,] #subset for only plant samples
alpha_sand_aov <- aov(evenness ~ Plastic, data=df)
alpha_table_Plastic_sand <- pairwise_test_Plastic(alpha_sand_aov, df)
pander(alpha_table_Plastic_sand, caption = "Plastic effect in sand")

#root_sand
df <- alpha_diversity_DAT_root_sand_summary[alpha_diversity_DAT_root_sand_summary$Plant == T,]
alpha_root_sand_aov <- aov(evenness ~ Plastic, data=df)
alpha_table_Plastic_root_sand <- pairwise_test_Plastic(alpha_root_sand_aov, df)
pander(alpha_table_Plastic_root_sand, caption = "Plastic effect in root_sand")

#root_soil
df <- alpha_diversity_DAT_root_soil_summary[alpha_diversity_DAT_root_soil_summary$Plant == T,]
alpha_root_soil_aov <- aov(evenness ~ Plastic, data=df)
alpha_table_Plastic_root_soil <- pairwise_test_Plastic(alpha_root_soil_aov, df)
pander(alpha_table_Plastic_root_soil, caption = "Plastic effect in root_soil")

#soil
df <- alpha_diversity_DAT_soil_summary[alpha_diversity_DAT_soil_summary$Plant == T,]
alpha_soil_aov <- aov(evenness ~ Plastic, data=df)
alpha_table_Plastic_soil <- pairwise_test_Plastic(alpha_soil_aov, df)
pander(alpha_table_Plastic_soil, caption = "Plastic effect in soil")

```

```{r plot a-div Plastic evenness, echo=F, message=F, warning=F, eval=T, , fig.height=5, fig.width=8}

#combine data.frames
alpha_diversity_DAT_summary <- rbind(alpha_diversity_DAT_sand_summary, alpha_diversity_DAT_root_sand_summary,
                                     alpha_diversity_DAT_root_soil_summary, alpha_diversity_DAT_soil_summary)
df <- alpha_diversity_DAT_summary[alpha_diversity_DAT_summary$Plant == T,]

#data frame for letters
alpha_table_Plastic <- rbind(alpha_table_Plastic_sand, alpha_table_Plastic_root_sand, alpha_table_Plastic_root_soil, alpha_table_Plastic_soil)
alpha_table_Plastic$Substrate <- rep(c("Sand", "Root_sand", "Root_soil", "Soil"), each=4)
alpha_table_Plastic$Substrate <- factor(alpha_table_Plastic$Substrate, levels=c("Sand", "Root_sand", "Root_soil", "Soil"))
alpha_table_Plastic$letters <- trimws(alpha_table_Plastic$letters) #remove whitespaces
alpha_table_Plastic$x <- rep(1:4, 4)
alpha_table_Plastic$y <- c(
tapply(alpha_diversity_DAT_sand_summary$evenness, alpha_diversity_DAT_sand_summary$Plastic, max)+0.02,
tapply(alpha_diversity_DAT_root_sand_summary$evenness, alpha_diversity_DAT_root_sand_summary$Plastic, max)+0.02,
tapply(alpha_diversity_DAT_root_soil_summary$evenness, alpha_diversity_DAT_root_soil_summary$Plastic, max)+0.02,
tapply(alpha_diversity_DAT_soil_summary$evenness, alpha_diversity_DAT_soil_summary$Plastic, max)+0.02)

#concentration
df$Concentration <- factor(paste0(df$Concentration, "%"), levels=c("0%", "3%", "10%"))

#plot
 
plot <- ggplot(df, aes(y=evenness, x=Plastic, fill=Plastic)) +
            geom_boxplot(position=position_dodge2(width=0.75, preserve="single"),outlier.colour = NA) +
            geom_jitter(size=1, position=position_jitterdodge(jitter.width=0.3,dodge.width = 0.75), alpha=0.7, aes(col=Concentration))+
            theme_bw() +
            theme(legend.position="none") +
            ylab("Pielou’s evenness")+
            xlab("")+
            scale_fill_manual(values=level_cols_Plastic) +
            scale_color_manual(values=c("black","dimgrey","grey")) +
            facet_grid(~ Substrate)+
            theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size=7), strip.text = element_text(size = 6)) +
            labs(title = "Bacteria: Alpha diversity", subtitle = "alpha diversity ~ Plastic + Plant")+
            geom_text(data = alpha_table_Plastic, aes(x=x, y=y, label=letters),size=4, col="dimgrey")



### PLOT

#combine the two plots (library cowplot)
fig4 <- plot_grid(plot + theme(legend.position="bottom"))

#print
#png("figures/fig4.3.png",6400, 4000, res = 600)
fig4
#dev.off()

```

\pagebreak

## Concentrations effect

**Q2: Are there differences in alpha diversity between the different concentrations?**\
Plants were either treated with no plastic (control), microplastic, macroplastic or starch. Plastic was added in a 3% or 10% concenctration. From the stats above we could see that the concentration has an effect on the alpha diversity in sand. We investigate now in which direction is the alpha diversity shifted by the different concentrations.

```{r a-div Concentration, warning=F, echo=F, message=F, eval=T}

pairwise_test_Concentration <- function(aov_data, summary_data){
  alpha_aov_tukey <- emmeans(aov_data, c("Concentration"))  # library(emmeans)
  alpha_aov_tukey_letters <- cld(alpha_aov_tukey, Letter="abcdefghi", alpha=0.05)
  # means
  alpha_means <- aggregate(summary_data$asv_diversity,list(summary_data$Concentration), mean)
  # table
  alpha_aov_tukey_letters <- alpha_aov_tukey_letters[order(with(alpha_aov_tukey_letters, Concentration)),]
  alpha_means <- alpha_means[order(with(alpha_means, Group.1, Group.2)),]
  alpha_table <- data.frame(Concentration=alpha_means[,1], Shannon=alpha_means[,2], letters=alpha_aov_tukey_letters[,7])
  return(alpha_table)
}

#sand
df <- alpha_diversity_DAT_sand_summary[alpha_diversity_DAT_sand_summary$Plant == T,] #subset for only plant samples
alpha_sand_aov <- aov(asv_diversity ~ Concentration, data=df)
alpha_table_Concentration_sand <- pairwise_test_Concentration(alpha_sand_aov, df)
alpha_table_Concentration_sand$Concentration <- paste0(alpha_table_Concentration_sand$Concentration, "%")
alpha_table_Concentration_sand <- alpha_table_Concentration_sand[c(1,3,2),]
pander(alpha_table_Concentration_sand, caption = "Concentration effect in sand")
#root_sand
df <- alpha_diversity_DAT_root_sand_summary[alpha_diversity_DAT_root_sand_summary$Plant == T,]
alpha_root_sand_aov <- aov(asv_diversity ~ Concentration, data=df)
alpha_table_Concentration_root_sand <- pairwise_test_Concentration(alpha_root_sand_aov, df)
alpha_table_Concentration_root_sand$Concentration <- paste0(alpha_table_Concentration_root_sand$Concentration, "%")
alpha_table_Concentration_root_sand <- alpha_table_Concentration_root_sand[c(1,3,2),]
pander(alpha_table_Concentration_root_sand, caption = "Concentration effect in root_sand")

#root_soil
df <- alpha_diversity_DAT_root_soil_summary[alpha_diversity_DAT_root_soil_summary$Plant == T,]
alpha_root_soil_aov <- aov(asv_diversity ~ Concentration, data=df)
alpha_table_Concentration_root_soil <- pairwise_test_Concentration(alpha_root_soil_aov, df)
alpha_table_Concentration_root_soil$Concentration <- paste0(alpha_table_Concentration_root_soil$Concentration, "%")
alpha_table_Concentration_root_soil <- alpha_table_Concentration_root_soil[c(1,3,2),]
pander(alpha_table_Concentration_root_soil, caption = "Concentration effect in root_soil")

#soil
df <- alpha_diversity_DAT_soil_summary[alpha_diversity_DAT_soil_summary$Plant == T,]
alpha_soil_aov <- aov(asv_diversity ~ Concentration, data=df)
alpha_table_Concentration_soil <- pairwise_test_Concentration(alpha_soil_aov, df)
alpha_table_Concentration_soil$Concentration <- paste0(alpha_table_Concentration_soil$Concentration, "%")
alpha_table_Concentration_soil <- alpha_table_Concentration_soil[c(1,3,2),]
pander(alpha_table_Concentration_soil, caption = "Concentration effect in soil")

```

### Figure 4.1 \| Concentration effect on alpha diversity
\vspace{5mm}

```{r plot a-div Concentration, echo=F, message=F, warning=F, eval=T, , fig.height=5, fig.width=8}

#combine data.frames
alpha_diversity_DAT_summary <- rbind(alpha_diversity_DAT_sand_summary, alpha_diversity_DAT_root_sand_summary,
                                     alpha_diversity_DAT_root_soil_summary, alpha_diversity_DAT_soil_summary)
df <- alpha_diversity_DAT_summary[alpha_diversity_DAT_summary$Plant == T,]

#data frame for letters
alpha_table_Concentration <- rbind(alpha_table_Concentration_sand, alpha_table_Concentration_root_sand, alpha_table_Concentration_root_soil, alpha_table_Concentration_soil)
alpha_table_Concentration$Substrate <- rep(c("Sand", "Root_sand", "Root_soil", "Soil"), each=3)
alpha_table_Concentration$Substrate <- factor(alpha_table_Concentration$Substrate, levels=c("Sand", "Root_sand", "Root_soil", "Soil"))
alpha_table_Concentration$letters <- trimws(alpha_table_Concentration$letters) #remove whitespaces
alpha_table_Concentration$x <- rep(c(1,2,3), 4)
alpha_table_Concentration$y <- c(
tapply(alpha_diversity_DAT_sand_summary$asv_diversity, alpha_diversity_DAT_sand_summary$Concentration, max)[c(1,3,2)]+15,
tapply(alpha_diversity_DAT_root_sand_summary$asv_diversity, alpha_diversity_DAT_root_sand_summary$Concentration, max)[c(1,3,2)]+15,
tapply(alpha_diversity_DAT_root_soil_summary$asv_diversity, alpha_diversity_DAT_root_soil_summary$Concentration, max)[c(1,3,2)]+15,
tapply(alpha_diversity_DAT_soil_summary$asv_diversity, alpha_diversity_DAT_soil_summary$Concentration, max)[c(1,3,2)]+15)

#concentration
df$Substrate <- factor(df$Substrate, levels=c("Sand", "Root_sand", "Root_soil", "Soil"))
df$Concentration <- factor(paste0(df$Concentration, "%"), levels=c("0%", "3%", "10%"))

#plot
 
plot <- ggplot(df, aes(y=asv_diversity, x=Concentration, fill=Concentration)) +
            geom_boxplot(position=position_dodge2(width=0.75, preserve="single"),outlier.colour = NA, alpha=0.7) +
            geom_jitter(size=1, position=position_jitterdodge(jitter.width=0.3,dodge.width = 0.75), alpha=0.7, aes(col=Plastic))+
            theme_bw() +
            theme(legend.position="none") +
            ylab("Shanon Diversity")+
            xlab("")+
            scale_fill_manual(values=c("black","dimgrey","grey")) +
            scale_color_manual(values=level_cols_Plastic) +
            facet_grid(~ Substrate)+
            theme(axis.text.x = element_text(angle = 90, vjust = 0.5, size=7), strip.text = element_text(size = 6)) +
            labs(title = "Bacteria: Alpha diversity", subtitle = "alpha diversity ~ Concentration + Plant")+
            geom_text(data = alpha_table_Concentration, aes(x=x, y=y, label=letters),size=4, col="dimgrey")



### PLOT

#combine the two plots (library cowplot)
fig4 <- plot_grid(plot + theme(legend.position="bottom"))

#print
#png("figures/fig4.png",6400, 4000, res = 600)
fig4
#dev.off()

```

\vspace{5mm}
**Conclusion:** A sand and soil is the alpha diversity is in overall reduced by treating with plastics. In roots from sand, the opposite effect can be observed. Roots from soil don't show effects. There is no clear effect that a higher concentration of plastic has a stronger effect than a lower concentration.

**Conclusion alpha diversity:** *In sand and soil, the alpha diversity was reduced by treating samples with microplastic, macroplastic or starch. In roots from sand or soil, the opposite effect could be observed. There is no clear effect, that a higher plastic concentration shows a stronger response on the alpha diversity. *

```{r export RDA files, , echo=F, message=F, warning=F}

# create directory
dir.create("interim")
dir.create("interim/03_Alpha_Diversity")

## set output directory
setwd("interim/03_Alpha_Diversity")

#save objects needed in the following scripts as RDA
saveRDS(alpha_diversity_DAT_summary, "alpha_diversity_DAT_summary.RDS")

```

\pagebreak
