---
title: "Microplastic - 05: Differential Abundance Analysis"
author: "Jan Waelchli"
geometry: margin=2cm
output:
  pdf_document:
    toc: yes
    toc_depth: 3
---
   
```{r setup, include=FALSE, echo=F, warning=F, message=F,}

##clear the object from memory
rm(list=ls())

#knitr settings
knitr::opts_chunk$set(echo=TRUE, fig.align="center")
options(tinytex.verbose = TRUE)

#set seed
set.seed(100)

#paths
paths <- list(functions="functions/")

## set source
library(rstudioapi)
setwd(dirname(getActiveDocumentContext()$path))

## load functions
source(paste0(paths$functions,"functions.R"))
functions(path=paths$functions)

## installs (if necessary) and loads libraries
libraries()
library(Biostrings)
library(ampir)
library(ggtree)
library(ggtreeExtra)
library(ggnewscale)

#set to true if tree-files were computed
treefiles <- T


```

```{r paths, echo=F, message=F, warning=F}

# Import

#set wd to RDS files
setwd("interim/01_Import_Normalization")

#Import RDS files
DESIGN <- readRDS("DESIGN.RDS")
DAT <- readRDS("DAT.RDS")
DAT_rare <- readRDS("DAT_rare.RDS")
TAX <- readRDS("TAX.RDS")

DESIGN_sand <- readRDS("DESIGN_sand.RDS")
DAT_sand <- readRDS("DAT_sand.RDS")
DAT_sand_rare <- readRDS("DAT_sand_rare.RDS")
TAX_sand <- readRDS("TAX_sand.RDS")

DESIGN_root_sand <- readRDS("DESIGN_root_sand.RDS")
DAT_root_sand <- readRDS("DAT_root_sand.RDS")
DAT_root_sand_rare <- readRDS("DAT_root_sand_rare.RDS")
TAX_root_sand <- readRDS("TAX_root_sand.RDS")

DESIGN_root_soil <- readRDS("DESIGN_root_soil.RDS")
DAT_root_soil <- readRDS("DAT_root_soil.RDS")
DAT_root_soil_rare <- readRDS("DAT_root_soil_rare.RDS")
TAX_root_soil <- readRDS("TAX_root_soil.RDS")

DESIGN_soil <- readRDS("DESIGN_soil.RDS")
DAT_soil <- readRDS("DAT_soil.RDS")
DAT_soil_rare <- readRDS("DAT_soil_rare.RDS")
TAX_soil <- readRDS("TAX_soil.RDS")

level_cols_Plastic <- readRDS("level_cols_Plastic.RDS")
level_cols_Substrate <- readRDS("level_cols_Substrate.RDS")
TAX_level_cols_phyla <- readRDS("TAX_level_cols_phyla.RDS")

PHYSEQ_phyla_melt <- readRDS("../02_Taxa_Analysis/PHYSEQ_phyla_melt.RDS")

```


\pagebreak

# Taxa Response

Is there a core of microbial taxa respond to the different plastic treatments? We searched sensitive ASVs – ASVs being differential abundant between CTRL samples and samples containing plastic.  We followed the same structure as before and answer the question:

**Q1: Are there sensitive ASVs between control and plastic-treated samples?**

We answer the question by using for different tools to measure differential abundances - aldex2, acombc, maaslin2 and metagenomeSeq - and predict ASVs to be different if they were detected by 3 or more tools.

```{r prepro, echo=F, message=F, warning=F}

preprocess <- function(DESIGN, DAT, sample_var){
  
  #preprocess (from ancom.R function file)
  #out_cut: detect outlayers
  #zero_cut: removes ASVs must be present in 90% of samples
  prepro <- feature_table_pre_process(feature_table=DAT, meta_data=DESIGN, 
                                       sample_var=sample_var, group_var=NULL, 
                                       out_cut=0.5, zero_cut=0.9,
                                       lib_cut=0, neg_lb=F)
  #overwrite
  feature_table <-  prepro$feature_table # Preprocessed feature table
  meta_data <-  prepro$meta_data # Preprocessed metadata
  return(list(feature_table, meta_data))
}

```

## Relative abundance in sand

We check for each ASV if it is sensitive or not to the different plastics. We show how many sensitive ASVs we have found for each plastic-treatment. Then, we plot the top 50 most abundant ASVs and label significant differences in red. Down below we show the phyla and class of those top abundant sensitive ASVs. 

```{r sand: compute DAA, echo=F, message=F, warning=F}

#microplastic

#subset to microplastic
DESIGN_sand_microplastic <-  DESIGN_sand[(DESIGN_sand$Plastic %in% c("CTRL", "microplastic") & DESIGN_sand$Plant==T),]
DAT_sand_microplastic <- DAT_sand[colnames(DAT_sand) %in% DESIGN_sand_microplastic$AccessArray_barcode] #for DAA analysis

#preprocess
temp <- preprocess(DESIGN_sand_microplastic, DAT_sand_microplastic, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_sand_microplastic <- temp[[1]]; rm(temp)
sens_ASVs_sand_microplastic <- summ_sand_microplastic$ASV[summ_sand_microplastic$score>2]


#macroplastic

#subset to macroplastic
DESIGN_sand_macroplastic <-  DESIGN_sand[(DESIGN_sand$Plastic %in% c("CTRL", "macroplastic") & DESIGN_sand$Plant==T),]
DAT_sand_macroplastic <- DAT_sand[colnames(DAT_sand) %in% DESIGN_sand_macroplastic$AccessArray_barcode] #forr DAA analysis

#preprocess
temp <- preprocess(DESIGN_sand_macroplastic, DAT_sand_macroplastic, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_sand_macroplastic <- temp[[1]]; rm(temp)
sens_ASVs_sand_macroplastic <- summ_sand_macroplastic$ASV[summ_sand_macroplastic$score>2]


#starch

#subset to starch
DESIGN_sand_starch <-  DESIGN_sand[(DESIGN_sand$Plastic %in% c("CTRL", "starch") & DESIGN_sand$Plant==T),]
DAT_sand_starch <- DAT_sand[colnames(DAT_sand) %in% DESIGN_sand_starch$AccessArray_barcode] #forr DAA analysis

#preprocess
temp <- preprocess(DESIGN_sand_starch, DAT_sand_starch, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_sand_starch <- temp[[1]]; rm(temp)
sens_ASVs_sand_starch <- summ_sand_starch$ASV[summ_sand_starch$score>2]

```


```{r sand: summarise DAA, echo=F, message=F, warning=F}

#relative abundance
DAT_sand_rare <- as.data.frame(DAT_sand_rare)
DAT_sand_rare <- (DAT_sand_rare/colSums(DAT_sand_rare)) * 100
DAT_sand_rare <- as.data.frame(DAT_sand_rare[ rowSums(DAT_sand_rare) > 0,])

#means & standard error
means <- aggregate(t(DAT_sand_rare), list(DESIGN_sand$Plastic), mean) %>%
  column_to_rownames("Group.1") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("ASV")

se <- aggregate(t(DAT_sand_rare), list(DESIGN_sand$Plastic), se) %>%
  column_to_rownames("Group.1") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("ASV")

#sort by abundance of CTRL samples
order <- order(-means$CTRL)
means <- means[order,]
se <- se[order,]

summary <- pivot_longer(means, !(ASV), names_to="Plastic", values_to="means")
summary$se <- pivot_longer(se, !(ASV), names_to="Plastic", values_to="se") %>% pull(se)
summary$Plastic <- factor(summary$Plastic, levels=c("CTRL", "microplastic", "macroplastic", "starch"))
summary$ASV <- factor(summary$ASV, levels=unique(summary$ASV)) #keep the sorted order of ASVs

#add score
#score
summ_sand_microplastic$plastic <- "microplastic"
summ_sand_macroplastic$plastic <- "macroplastic"
summ_sand_starch$plastic <- "starch"
summ <- rbind(summ_sand_microplastic, summ_sand_macroplastic, summ_sand_starch)

#get score
summary$score <- 0
for(p in unique(summary$Plastic)){
  for(a in unique(summary$ASV)){
    score <- summ %>% filter(plastic == p) %>% filter(ASV == a) %>% pull(score)
    if(length(score)==1){summary$score[(summary$Plastic==p & summary$ASV==a)] <- score}
  }}

#add if different abundant (score > 2)
summary$sensitive <- ifelse(summary$score > 2, "sensitive", "not_sensitive")
summary$sensitive <- as.factor(summary$sensitive)

#add abundace in trt compared to ctrl
summary$abu_trt <- "unchanged"
for(p in unique(summary$Plastic)){
  for(a in unique(summary$ASV)){
    mean_ctrl <- summary$means[(summary$ASV==a) & (summary$Plastic=="CTRL")]
    mean_trt <- summary$means[(summary$ASV==a) & (summary$Plastic==p)]
    abu_sens <- summary$sensitive[(summary$ASV==a) & (summary$Plastic==p)]
    if(abu_sens == "sensitive"){
      trt_dir <- ifelse(mean_ctrl > mean_trt, "higher in CTRL", "lower in CTRL")
      summary$abu_trt[(summary$ASV==a) & (summary$Plastic==p)] <- trt_dir
      }
  }
}

#add phyla and class
summary$phylum <- NA
summary$class <- NA
summary$order <- NA
summary$family <- NA
summary$genus <- NA
for(a in unique(summary$ASV)){
  if(a %in% TAX$ASV_ID){
    summary$phylum[summary$ASV == a] <- as.character(TAX$phylum[TAX$ASV_ID == a])
    summary$class[summary$ASV == a] <- as.character(TAX$class[TAX$ASV_ID == a])
    summary$order[summary$ASV == a] <- as.character(TAX$order[TAX$ASV_ID == a])
    summary$family[summary$ASV == a] <- as.character(TAX$family[TAX$ASV_ID == a])
    summary$genus[summary$ASV == a] <- as.character(TAX$genus[TAX$ASV_ID == a])
  }
  else{
    summary$phylum[summary$ASV == a] <- "unassigned"
    summary$class[summary$ASV == a] <- "unassigned"
    summary$order[summary$ASV == a] <- "unassigned"
    summary$family[summary$ASV == a] <- "unassigned"
    summary$genus[summary$ASV == a] <- "unassigned"
  }
}

summary_sand <- summary

```

### Different abundant ASVs

We show how many ASVs has been changed due to the plastic treatments how much of the relative abundance belongs to those sensitive ASVs.

```{r sand: show DAA microplastic, echo=F, message=F, warning=F}

#pander how many ASVs are sensitive

#microplastic
summary_sand_microplastic <- summary[(summary$Plastic %in% c("CTRL", "microplastic")) & (summary$ASV %in% summ_sand_microplastic$ASV),]
changed_abu_sand_microplastic <- sum(summary_sand_microplastic$means[summary_sand_microplastic$Plastic=="microplastic" &
                                      summary_sand_microplastic$sensitive=="sensitive"])
changed_abu_sand_microplastic <- paste0(round(changed_abu_sand_microplastic,2), "%")
sens_sand_microplastic <- table(summary_sand_microplastic$abu_trt[summary_sand_microplastic$Plastic == "microplastic"])
pander(sens_sand_microplastic[c(2,3,1)], caption = "microplastic effect in sand")

```

`r changed_abu_sand_microplastic` of the bacterial community in sand was changed in abundance due to microplastic.

```{r sand: show DAA macroplastic, echo=F, message=F, warning=F}

#macroplastic
summary_sand_macroplastic <- summary[(summary$Plastic %in% c("CTRL", "macroplastic")) & (summary$ASV %in% summ_sand_macroplastic$ASV),]
changed_abu_sand_macroplastic <- sum(summary_sand_macroplastic$means[summary_sand_macroplastic$Plastic=="macroplastic" &
                                      summary_sand_macroplastic$sensitive=="sensitive"])
changed_abu_sand_macroplastic <- paste0(round(changed_abu_sand_macroplastic,2), "%")
sens_sand_macroplastic <- table(summary_sand_macroplastic$abu_trt[summary_sand_macroplastic$Plastic == "macroplastic"])
pander(sens_sand_macroplastic[c(2,3,1)], caption = "macroplastic effect in sand")

```

`r changed_abu_sand_macroplastic` of the bacterial community in sand was changed in abundance due to macroplastic.

```{r sand: show DAA starch, echo=F, message=F, warning=F}

#starch
summary_sand_starch <- summary[(summary$Plastic %in% c("CTRL", "starch")) & (summary$ASV %in% summ_sand_starch$ASV),]
changed_abu_sand_starch <- sum(summary_sand_starch$means[summary_sand_starch$Plastic=="starch" &
                                      summary_sand_starch$sensitive=="sensitive"])
changed_abu_sand_starch <- paste0(round(changed_abu_sand_starch,2), "%")
sens_sand_starch <- table(summary_sand_starch$abu_trt[summary_sand_starch$Plastic == "starch"])
pander(sens_sand_starch[c(2,3,1)], caption = "starch effect in sand")

```

`r changed_abu_sand_starch` of the bacterial community in sand was changed in abundance due to starch.

\pagebreak

### Relative abundance plot

We plot the 50 top abundant ASVs (in controls) and label plastic sensitive ASVs in red.

\vspace{5mm}

```{r sand: relative abundance plot, echo=F, message=F, warning=F}

#plot the 50 most abundant ASVs

#set threshold for top abundant
top_abu <- 50

#plot
ggplot(summary[1:(top_abu*4),], aes(ASV, means, fill=Plastic, col=sensitive))+
  geom_bar(stat="identity", position="dodge", width = 0.8, size=0.2)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, size=6))+
  scale_color_manual(values=c(NA, "red"))+
  scale_fill_manual(values=level_cols_Plastic)+
  geom_errorbar(aes(ymin=means, ymax=means+se/2), width=.3, position=position_dodge(.8))
  #geom_text(aes(label = score, y=-0.2), position = position_dodge(.8), size=1)

```

\pagebreak

### Taxonomy of abundant ASVs

We show the taxonomy of top abundant ASVs which are plastic sensitive.

```{r sand: relative abundance taxa microplastic, echo=F, message=F, warning=F}

#microplastic

#sensitive ASVs from top 50
top_sensitive_microplastic <- pivot_wider(summary_sand_microplastic,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_microplastic <- top_sensitive_microplastic[1:top_abu,] #top abundant
top_sensitive_microplastic <- top_sensitive_microplastic[top_sensitive_microplastic$sensitive_microplastic =="sensitive",] #sensitive
top_sensitive_microplastic <- top_sensitive_microplastic[,colnames(top_sensitive_microplastic) %in%
                                                         c("ASV","means_CTRL","means_microplastic", "phylum", "class")] #column to pander
top_sensitive_microplastic$means_CTRL <- paste0(round(top_sensitive_microplastic$means_CTRL,2),"%")
top_sensitive_microplastic$means_microplastic <- paste0(round(top_sensitive_microplastic$means_microplastic,2),"%")
colnames(top_sensitive_microplastic)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_microplastic, caption="Top abundant ASVs sensitive to microplastic in sand")

```

\pagebreak

```{r sand: relative abundance taxa macroplastic, echo=F, message=F, warning=F}

#macroplastic

#sensitive ASVs from top 50
top_sensitive_macroplastic <- pivot_wider(summary_sand_macroplastic,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_macroplastic <- top_sensitive_macroplastic[1:top_abu,] #top abundant
top_sensitive_macroplastic <- top_sensitive_macroplastic[top_sensitive_macroplastic$sensitive_macroplastic =="sensitive",] #sensitive
top_sensitive_macroplastic <- top_sensitive_macroplastic[,colnames(top_sensitive_macroplastic) %in%
                                                         c("ASV","means_CTRL","means_macroplastic", "phylum", "class")] #column to pander
top_sensitive_macroplastic$means_CTRL <- paste0(round(top_sensitive_macroplastic$means_CTRL,2),"%")
top_sensitive_macroplastic$means_macroplastic <- paste0(round(top_sensitive_macroplastic$means_macroplastic,2),"%")
colnames(top_sensitive_macroplastic)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_macroplastic, caption="Top abundant ASVs sensitive to macroplastic in sand")

```

\pagebreak

```{r sand: relative abundance taxa starch, echo=F, message=F, warning=F}

#starch

#sensitive ASVs from top 50
top_sensitive_starch <- pivot_wider(summary_sand_starch,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_starch <- top_sensitive_starch[1:top_abu,] #top abundant
top_sensitive_starch <- top_sensitive_starch[top_sensitive_starch$sensitive_starch =="sensitive",] #sensitive
top_sensitive_starch <- top_sensitive_starch[,colnames(top_sensitive_starch) %in%
                                                         c("ASV","means_CTRL","means_starch", "phylum", "class")] #column to pander
top_sensitive_starch$means_CTRL <- paste0(round(top_sensitive_starch$means_CTRL,2),"%")
top_sensitive_starch$means_starch <- paste0(round(top_sensitive_starch$means_starch,2),"%")
colnames(top_sensitive_starch)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_starch, caption="Top abundant ASVs sensitive to starch in sand")

```

\pagebreak

### Sensitive Taxonomies

We plot how many ASVs are sensitive to multiple plastic treatments and show frequent phyla, classes and families which are sensitive to all three plastic treatments.

\vspace{5mm}

```{r sand: Venn plot, echo=F, message=F, warning=F}

ggvenn(list(microplastic=summary_sand_microplastic$ASV[summary_sand_microplastic$sensitive=="sensitive"],
            macroplastic=summary_sand_macroplastic$ASV[summary_sand_macroplastic$sensitive=="sensitive"],
            starch=summary_sand_starch$ASV[summary_sand_starch$sensitive=="sensitive"]),
  fill_color = c("darkolivegreen2", 'darkseagreen4', 'gold'),
  stroke_size = 0.5, set_name_size = 5, 
  #set_name_color = c("darkolivegreen2", 'darkseagreen4', 'gold'),
  show_percentage = F
  )+
  ggtitle("Sand: Sensitive bacterial ASVs in sand")

```

```{r sand: frequent sensitive taxa, echo=F, message=F, warning=F}

# #sensitive taxonomy frequent in microplastic
# 
# #phylum
# sensitive_microplastic_phylum <- summary_sand_microplastic[summary_sand_microplastic$sensitive=="sensitive",] %>% pull("phylum") %>% table() %>% sort(decreasing = T)
# sensitive_microplastic_phylum <- as.data.frame(sensitive_microplastic_phylum[sensitive_microplastic_phylum >= 5]) #filter by phylum with more than 4 sensitive ASVs
# colnames(sensitive_microplastic_phylum) <- c("phylum", "frequency")
# pander(sensitive_microplastic_phylum, caption="Frequent phyla sensitive to microplastic  in sand")
# 
# #class
# sensitive_microplastic_class <- summary_sand_microplastic[summary_sand_microplastic$sensitive=="sensitive",] %>% pull("class") %>% table() %>% sort(decreasing = T)
# sensitive_microplastic_class <- as.data.frame(sensitive_microplastic_class[sensitive_microplastic_class >= 5]) #filter by class with more than 4 sensitive ASVs
# colnames(sensitive_microplastic_class) <- c("class", "frequency")
# pander(sensitive_microplastic_class, caption="Frequent classes sensitive to microplastic  in sand")
# 
# #familiy
# sensitive_microplastic_family <- summary_sand_microplastic[summary_sand_microplastic$sensitive=="sensitive",] %>% pull("family") %>% table() %>% sort(decreasing = T)
# sensitive_microplastic_family <- as.data.frame(sensitive_microplastic_family[sensitive_microplastic_family >= 5]) #filter by family with more than 4 sensitive ASVs
# colnames(sensitive_microplastic_family) <- c("family", "frequency")
# pander(sensitive_microplastic_family, caption="Frequent families sensitive to microplastic  in sand")
# 
# 
# #sensitive taxonomy frequent in macroplastic
# 
# #phylum
# sensitive_macroplastic_phylum <- summary_sand_macroplastic[summary_sand_macroplastic$sensitive=="sensitive",] %>% pull("phylum") %>% table() %>% sort(decreasing = T)
# sensitive_macroplastic_phylum <- as.data.frame(sensitive_macroplastic_phylum[sensitive_macroplastic_phylum >= 5]) #filter by phylum with more than 4 sensitive ASVs
# colnames(sensitive_macroplastic_phylum) <- c("phylum", "frequency")
# pander(sensitive_macroplastic_phylum, caption="Frequent phyla sensitive to macroplastic  in sand")
# 
# #class
# sensitive_macroplastic_class <- summary_sand_macroplastic[summary_sand_macroplastic$sensitive=="sensitive",] %>% pull("class") %>% table() %>% sort(decreasing = T)
# sensitive_macroplastic_class <- as.data.frame(sensitive_macroplastic_class[sensitive_macroplastic_class >= 5]) #filter by class with more than 4 sensitive ASVs
# colnames(sensitive_macroplastic_class) <- c("class", "frequency")
# pander(sensitive_macroplastic_class, caption="Frequent classes sensitive to macroplastic  in sand")
# 
# #familiy
# sensitive_macroplastic_family <- summary_sand_macroplastic[summary_sand_macroplastic$sensitive=="sensitive",] %>% pull("family") %>% table() %>% sort(decreasing = T)
# sensitive_macroplastic_family <- as.data.frame(sensitive_macroplastic_family[sensitive_macroplastic_family >= 5]) #filter by family with more than 4 sensitive ASVs
# colnames(sensitive_macroplastic_family) <- c("family", "frequency")
# pander(sensitive_macroplastic_family, caption="Frequent families sensitive to macroplastic  in sand")
# 
# 
# #sensitive taxonomy frequent in starch
# 
# #phylum
# sensitive_starch_phylum <- summary_sand_starch[summary_sand_starch$sensitive=="sensitive",] %>% pull("phylum") %>% table() %>% sort(decreasing = T)
# sensitive_starch_phylum <- as.data.frame(sensitive_starch_phylum[sensitive_starch_phylum >= 5]) #filter by phylum with more than 4 sensitive ASVs
# colnames(sensitive_starch_phylum) <- c("phylum", "frequency")
# pander(sensitive_starch_phylum, caption="Frequent phyla sensitive to starch  in sand")
# 
# #class
# sensitive_starch_class <- summary_sand_starch[summary_sand_starch$sensitive=="sensitive",] %>% pull("class") %>% table() %>% sort(decreasing = T)
# sensitive_starch_class <- as.data.frame(sensitive_starch_class[sensitive_starch_class >= 5]) #filter by class with more than 4 sensitive ASVs
# colnames(sensitive_starch_class) <- c("class", "frequency")
# pander(sensitive_starch_class, caption="Frequent classes sensitive to starch  in sand")
# 
# #familiy
# sensitive_starch_family <- summary_sand_starch[summary_sand_starch$sensitive=="sensitive",] %>% pull("family") %>% table() %>% sort(decreasing = T)
# sensitive_starch_family <- as.data.frame(sensitive_starch_family[sensitive_starch_family >= 5]) #filter by family with more than 4 sensitive ASVs
# colnames(sensitive_starch_family) <- c("family", "frequency")
# pander(sensitive_starch_family, caption="Frequent families sensitive to starch  in sand")

```

\pagebreak

```{r sand: common frequent sensitive taxa, echo=F, message=F, warning=F}

#sensitive taxonomy frequent in microplastic, macroplastic and starch
sensitive_common_phylum <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV, family), names_from = "Plastic", values_from = c("phylum")) %>%
                             na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
colnames(sensitive_common_phylum)[2] <- "phylum"
sensitive_common_phylum <- as.data.frame(sort(table(sensitive_common_phylum$phylum),decreasing = T)) %>% filter(Freq >= 5) #filter by phylum with more than 4 sensitive ASVs
colnames(sensitive_common_phylum) <- c("phylum", "frequency")
pander(sensitive_common_phylum, caption="Frequent phyla sensitive to all plastic treatments  in sand")

#class
sensitive_common_class <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV, family), names_from = "Plastic", values_from = c("class")) %>%
                             na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
colnames(sensitive_common_class)[2] <- "class"
sensitive_common_class <- as.data.frame(sort(table(sensitive_common_class$class),decreasing = T)) %>% filter(Freq >= 5) #filter by class with more than 4 sensitive ASVs
colnames(sensitive_common_class) <- c("class", "frequency")
pander(sensitive_common_class, caption="Frequent classes sensitive to all plastic treatments in sand")

#family
sensitive_common_family <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV), names_from = "Plastic", values_from = c("family")) %>%
                             na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
colnames(sensitive_common_family)[2] <- "family"
sensitive_common_family <- as.data.frame(sort(table(sensitive_common_family$family),decreasing = T)) %>% filter(Freq >= 5) #filter by family with more than 4 sensitive ASVs
colnames(sensitive_common_family) <- c("family", "frequency")
pander(sensitive_common_family, caption="Frequent families sensitive to all plastic treatments in sand")

```

\pagebreak

### Phylogenetic tree

We plot a phylogenetic tree with all ASVs tested for sensitivity and label with the rings around the tree if they are sensitive to microplastic, macroplastic and starch.

\vspace{5mm}

```{r sand: subset sequences for tree, echo=F, message=F, warning=F}

#import seqs
seqs <- Biostrings::readDNAStringSet("../01_dadapipe/4_output/bacteria_ASV/ASV100/bacteria_SEQ100.fasta")
seqs <- as.data.frame(seqs); colnames(seqs) <- "seqs"

#subset for tree
tested_ASVs_sand <- unique(summ$ASV)
seqs_sand <- seqs[rownames(seqs) %in% tested_ASVs_sand,,drop=F]
seqs_sand <- data.frame(name=rownames(seqs_sand), seq=seqs_sand$seqs)
rownames(seqs_sand) <- seqs_sand$name

#export as fasta file
#dir.create("../03_build_tree/sand",recursive = T)
#ampir::df_to_faa(seqs_sand, "../03_build_tree/sand/sequences_sand.fasta")

########
#in bash
#cd ../03_build_tree/sand/sequences_sand
#mafft-linsi sequences_sand.fasta > sequences_sand_aligned.fasta #align sequences
#iqtree -s sequences_sand_aligned.fasta #build tree
########
```

```{r sand: show taxa tree, echo=F, message=F, warning=F}

#set to true at the beginning of script if files were computed
if(treefiles){

  #read tree
  tree_file <- "/Users/j.waechli/Desktop/Jan_Waelchli/01_Projects/01_Support/Robin/Microbiome_BE13/03_build_tree/sand/sequences_sand_aligned.fasta.treefile"
  tree <- read.tree(tree_file)
  
  #subset summary and sort them
  summary_taxa <- unique(summary[,c("ASV", "phylum","class","family")])
  summary_taxa <- summary_taxa[summary_taxa$ASV %in% tested_ASVs_sand,]
  rownames(summary_taxa) <- summary_taxa$ASV
  summary_taxa <- summary_taxa[tree[["tip.label"]],]
  
  
  #colorstring
  summary_taxa$labels <- NA
  summary_taxa$cols <- NA
  
  for(a in summary_taxa$ASV){
    summary_taxa$labels[summary_taxa$ASV == a] <- as.character(TAX$labels[TAX$ASV_ID == a][1])
    summary_taxa$cols[summary_taxa$ASV == a] <- as.character(TAX$cols_phyla[TAX$ASV_ID == a][1])
  }
  
  high_abundant_phyla <- as.character(unique(PHYSEQ_phyla_melt$labels_2))
  summary_taxa$labels_2 <- summary_taxa$labels
  summary_taxa$labels_2[!(summary_taxa$labels %in% high_abundant_phyla)] <- "Low abundant phyla"
  summary_taxa$cols2 <- summary_taxa$cols
  summary_taxa$cols2[summary_taxa$labels_2 == "Low abundant phyla"] <- "lightgrey"
  
  #extend taxa string to the length of tree edges (needs for each edge a value but uses only the number of tips)
  phylum <- c(summary_taxa$labels_2,rep(NA,(length(tree[["edge.length"]])-length(summary_taxa$labels_2))+1))
  cols2 <- summary_taxa$cols2
  names(cols2) <- summary_taxa$labels_2
  cols2_long <- c(cols2,rep("white",(length(tree[["edge.length"]])-length(cols2))+1))
  
  #plot genetic tree with phyla
  p1 <- ggtree(tree, layout = "fan", branch.length = "none")+
    geom_tippoint(aes(color=phylum), size=2)+
    scale_color_manual(values=cols2_long)+
    new_scale_color()
  
  
  #for saving
  phylotree_sand <- list(plot=p1, phylum=phylum, colstring=cols2_long)
  
  
  #plot microplastic ring
  df <- summary[summary$Plastic %in% c("CTRL", "microplastic") & summary$ASV %in% rownames(seqs_sand),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_microplastic) / (df$means_CTRL + df$means_microplastic)) *100
  df$abu_trt_microplastic <- factor(df$abu_trt_microplastic, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p2 <- p1 + geom_rect(aes(xmin = 78, xmax = 115, ymin = 1, ymax = nrow(df)), fill="darkolivegreen2")+ #background color
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_microplastic),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list()) +
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  #plot macroplastic ring
  df <- summary[summary$Plastic %in% c("CTRL", "macroplastic") & summary$ASV %in% rownames(seqs_sand),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_macroplastic) / (df$means_CTRL + df$means_macroplastic)) *100
  df$abu_trt_macroplastic <- factor(df$abu_trt_macroplastic, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p3 <- p2 + geom_rect(aes(xmin = 115, xmax = 155, ymin = 1, ymax = nrow(df)), fill="darkseagreen4")+
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_macroplastic),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list())+
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  
  #plot starch ring
  df <- summary[summary$Plastic %in% c("CTRL", "starch") & summary$ASV %in% rownames(seqs_sand),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_starch) / (df$means_CTRL + df$means_starch)) *100
  df$abu_trt_starch <- factor(df$abu_trt_starch, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p4 <- p3 + geom_rect(aes(xmin = 155, xmax = 188, ymin = 1, ymax = nrow(df)), fill="gold")+
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_starch),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list())+
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  #pseudo plot for legend
  library(ggnewscale)
  df <- data.frame(Phylum=names(cols2), Plastic=c(rep("microplastic",length(cols2)-2),"macroplastic","starch"),
                   Sensitivity=c(rep("higher in CTRL",length(cols2)-2),"unchanged","lower in CTRL"))
  df$Sensitivity <- factor(df$Sensitivity, levels=c("higher in CTRL", "unchanged","lower in CTRL"))
  legend_plot <- ggplot(df,aes(1,1))+
                    geom_bar(stat="identity", aes(fill=Plastic))+
                    geom_point(pch=15, size=6, aes(col=Phylum))+
                    scale_color_manual(values=cols2)+
                    scale_fill_manual(values=level_cols_Plastic[c(1,2,4)], name = "Plastic")+
                    theme_bw()+
                    new_scale_color() +
                    geom_point(pch=15, size=6, aes(col=Sensitivity))+
                    scale_color_manual(values=c("tomato", "grey80", "darkcyan"))
                    #guides(size = guide_legend(order = 1), color = guide_legend(order = 2))
  
  #add plot and legend
  phyla_plot <- plot_grid(p4 + theme(legend.position="none"), 
                          get_legend(legend_plot), nrow=1,rel_widths=c(1, .3))
  phyla_plot
  
}

```

\vspace{5mm}

**Conclusion differential abundance analysis:** *A huge part of the microbiome is shaped by adding plastic to the soil. Microplastic, macroplastic and starch has a similar effect. Bacteria reacting to the plastic are phylogenetic diverse.*

\pagebreak

## Relative abundance in root sand

We check for each ASV if it is sensitive or not to the different plastics. We show how many sensitive ASVs we have found for each plastic-treatment. Then, we plot the top 50 most abundant ASVs and label significant differences in red. Down below we show the phyla and class of those top abundant sensitive ASVs. 

```{r root_sand: compute DAA, echo=F, message=F, warning=F}

#microplastic

#subset to microplastic
DESIGN_root_sand_microplastic <-  DESIGN_root_sand[(DESIGN_root_sand$Plastic %in% c("CTRL", "microplastic") & DESIGN_root_sand$Plant==T),]
DAT_root_sand_microplastic <- DAT_root_sand[colnames(DAT_root_sand) %in% DESIGN_root_sand_microplastic$AccessArray_barcode] #for DAA analysis

#preprocess
temp <- preprocess(DESIGN_root_sand_microplastic, DAT_root_sand_microplastic, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_root_sand_microplastic <- temp[[1]]; rm(temp)
sens_ASVs_root_sand_microplastic <- summ_root_sand_microplastic$ASV[summ_root_sand_microplastic$score>2]


#macroplastic

#subset to macroplastic
DESIGN_root_sand_macroplastic <-  DESIGN_root_sand[(DESIGN_root_sand$Plastic %in% c("CTRL", "macroplastic") & DESIGN_root_sand$Plant==T),]
DAT_root_sand_macroplastic <- DAT_root_sand[colnames(DAT_root_sand) %in% DESIGN_root_sand_macroplastic$AccessArray_barcode] #forr DAA analysis

#preprocess
temp <- preprocess(DESIGN_root_sand_macroplastic, DAT_root_sand_macroplastic, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_root_sand_macroplastic <- temp[[1]]; rm(temp)
sens_ASVs_root_sand_macroplastic <- summ_root_sand_macroplastic$ASV[summ_root_sand_macroplastic$score>2]


#starch

#subset to starch
DESIGN_root_sand_starch <-  DESIGN_root_sand[(DESIGN_root_sand$Plastic %in% c("CTRL", "starch") & DESIGN_root_sand$Plant==T),]
DAT_root_sand_starch <- DAT_root_sand[colnames(DAT_root_sand) %in% DESIGN_root_sand_starch$AccessArray_barcode] #forr DAA analysis

#preprocess
temp <- preprocess(DESIGN_root_sand_starch, DAT_root_sand_starch, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_root_sand_starch <- temp[[1]]; rm(temp)
sens_ASVs_root_sand_starch <- summ_root_sand_starch$ASV[summ_root_sand_starch$score>2]

```


```{r root_sand: summarise DAA, echo=F, message=F, warning=F}

#relative abundance
DAT_root_sand_rare <- as.data.frame(DAT_root_sand_rare)
DAT_root_sand_rare <- (DAT_root_sand_rare/colSums(DAT_root_sand_rare)) * 100
DAT_root_sand_rare <- as.data.frame(DAT_root_sand_rare[ rowSums(DAT_root_sand_rare) > 0,])

#means & standard error
means <- aggregate(t(DAT_root_sand_rare), list(DESIGN_root_sand$Plastic), mean) %>%
  column_to_rownames("Group.1") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("ASV")

se <- aggregate(t(DAT_root_sand_rare), list(DESIGN_root_sand$Plastic), se) %>%
  column_to_rownames("Group.1") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("ASV")

#sort by abundance of CTRL samples
order <- order(-means$CTRL)
means <- means[order,]
se <- se[order,]

summary <- pivot_longer(means, !(ASV), names_to="Plastic", values_to="means")
summary$se <- pivot_longer(se, !(ASV), names_to="Plastic", values_to="se") %>% pull(se)
summary$Plastic <- factor(summary$Plastic, levels=c("CTRL", "microplastic", "macroplastic", "starch"))
summary$ASV <- factor(summary$ASV, levels=unique(summary$ASV)) #keep the sorted order of ASVs

#add score
#score
summ_root_sand_microplastic$plastic <- "microplastic"
summ_root_sand_macroplastic$plastic <- "macroplastic"
summ_root_sand_starch$plastic <- "starch"
summ <- rbind(summ_root_sand_microplastic, summ_root_sand_macroplastic, summ_root_sand_starch)

#get score
summary$score <- 0
for(p in unique(summary$Plastic)){
  for(a in unique(summary$ASV)){
    score <- summ %>% filter(plastic == p) %>% filter(ASV == a) %>% pull(score)
    if(length(score)==1){summary$score[(summary$Plastic==p & summary$ASV==a)] <- score}
  }}

#add if different abundant (score > 2)
summary$sensitive <- ifelse(summary$score > 2, "sensitive", "not_sensitive")
summary$sensitive <- as.factor(summary$sensitive)

#add abundace in trt compared to ctrl
summary$abu_trt <- "unchanged"
for(p in unique(summary$Plastic)){
  for(a in unique(summary$ASV)){
    mean_ctrl <- summary$means[(summary$ASV==a) & (summary$Plastic=="CTRL")]
    mean_trt <- summary$means[(summary$ASV==a) & (summary$Plastic==p)]
    abu_sens <- summary$sensitive[(summary$ASV==a) & (summary$Plastic==p)]
    if(abu_sens == "sensitive"){
      trt_dir <- ifelse(mean_ctrl > mean_trt, "higher in CTRL", "lower in CTRL")
      summary$abu_trt[(summary$ASV==a) & (summary$Plastic==p)] <- trt_dir
      }
  }
}

#add phyla and class
summary$phylum <- NA
summary$class <- NA
summary$order <- NA
summary$family <- NA
summary$genus <- NA
for(a in unique(summary$ASV)){
  if(a %in% TAX$ASV_ID){
    summary$phylum[summary$ASV == a] <- as.character(TAX$phylum[TAX$ASV_ID == a])
    summary$class[summary$ASV == a] <- as.character(TAX$class[TAX$ASV_ID == a])
    summary$order[summary$ASV == a] <- as.character(TAX$order[TAX$ASV_ID == a])
    summary$family[summary$ASV == a] <- as.character(TAX$family[TAX$ASV_ID == a])
    summary$genus[summary$ASV == a] <- as.character(TAX$genus[TAX$ASV_ID == a])
  }
  else{
    summary$phylum[summary$ASV == a] <- "unassigned"
    summary$class[summary$ASV == a] <- "unassigned"
    summary$order[summary$ASV == a] <- "unassigned"
    summary$family[summary$ASV == a] <- "unassigned"
    summary$genus[summary$ASV == a] <- "unassigned"
  }
}

summary_root_sand <- summary

```

### Different abundant ASVs

We show how many ASVs has been changed due to the plastic treatments how much of the relative abundance belongs to those sensitive ASVs.

```{r root_sand: show DAA microplastic, echo=F, message=F, warning=F}

#pander how many ASVs are sensitive

#microplastic
summary_root_sand_microplastic <- summary[(summary$Plastic %in% c("CTRL", "microplastic")) & (summary$ASV %in% summ_root_sand_microplastic$ASV),]
changed_abu_root_sand_microplastic <- sum(summary_root_sand_microplastic$means[summary_root_sand_microplastic$Plastic=="microplastic" &
                                      summary_root_sand_microplastic$sensitive=="sensitive"])
changed_abu_root_sand_microplastic <- paste0(round(changed_abu_root_sand_microplastic,2), "%")
sens_root_sand_microplastic <- table(summary_root_sand_microplastic$abu_trt[summary_root_sand_microplastic$Plastic == "microplastic"])
pander(sens_root_sand_microplastic[c(2,3,1)], caption = "microplastic effect in root_sand")

```

`r changed_abu_root_sand_microplastic` of the bacterial community in root was changed in abundance due to microplastic.

```{r root_sand: show DAA macroplastic, echo=F, message=F, warning=F}

#macroplastic
summary_root_sand_macroplastic <- summary[(summary$Plastic %in% c("CTRL", "macroplastic")) & (summary$ASV %in% summ_root_sand_macroplastic$ASV),]
changed_abu_root_sand_macroplastic <- sum(summary_root_sand_macroplastic$means[summary_root_sand_macroplastic$Plastic=="macroplastic" &
                                      summary_root_sand_macroplastic$sensitive=="sensitive"])
changed_abu_root_sand_macroplastic <- paste0(round(changed_abu_root_sand_macroplastic,2), "%")
sens_root_sand_macroplastic <- table(summary_root_sand_macroplastic$abu_trt[summary_root_sand_macroplastic$Plastic == "macroplastic"])
pander(sens_root_sand_macroplastic[c(2,3,1)], caption = "macroplastic effect in root_sand")

```

`r changed_abu_root_sand_macroplastic` of the bacterial community in root was changed in abundance due to macroplastic.

```{r root_sand: show DAA starch, echo=F, message=F, warning=F}

#starch
summary_root_sand_starch <- summary[(summary$Plastic %in% c("CTRL", "starch")) & (summary$ASV %in% summ_root_sand_starch$ASV),]
changed_abu_root_sand_starch <- sum(summary_root_sand_starch$means[summary_root_sand_starch$Plastic=="starch" &
                                      summary_root_sand_starch$sensitive=="sensitive"])
changed_abu_root_sand_starch <- paste0(round(changed_abu_root_sand_starch,2), "%")
sens_root_sand_starch <- table(summary_root_sand_starch$abu_trt[summary_root_sand_starch$Plastic == "starch"])
pander(sens_root_sand_starch[c(2,3,1)], caption = "starch effect in root_sand")

```

`r changed_abu_root_sand_starch` of the bacterial community in root_sand was changed in abundance due to starch.

\pagebreak

### Relative abundance plot

We plot the 50 top abundant ASVs (in controls) and label plastic sensitive ASVs in red.

\vspace{5mm}

```{r root_sand: relative abundance plot, echo=F, message=F, warning=F}

#plot the 50 most abundant ASVs

#set threshold for top abundant
top_abu <- 50

#plot
ggplot(summary[1:(top_abu*4),], aes(ASV, means, fill=Plastic, col=sensitive))+
  geom_bar(stat="identity", position="dodge", width = 0.8, size=0.2)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, size=6))+
  scale_color_manual(values=c(NA, "red"))+
  scale_fill_manual(values=level_cols_Plastic)+
  geom_errorbar(aes(ymin=means, ymax=means+se/2), width=.3, position=position_dodge(.8))
  #geom_text(aes(label = score, y=-0.2), position = position_dodge(.8), size=1)

```

\pagebreak

### Taxonomy of abundant ASVs

We show the taxonomy of top abundant ASVs which are plastic sensitive.

```{r root_sand: relative abundance taxa microplastic, echo=F, message=F, warning=F}

#microplastic

#sensitive ASVs from top 50
top_sensitive_microplastic <- pivot_wider(summary_root_sand_microplastic,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_microplastic <- top_sensitive_microplastic[1:top_abu,] #top abundant
top_sensitive_microplastic <- top_sensitive_microplastic[top_sensitive_microplastic$sensitive_microplastic =="sensitive",] #sensitive
top_sensitive_microplastic <- top_sensitive_microplastic[,colnames(top_sensitive_microplastic) %in%
                                                         c("ASV","means_CTRL","means_microplastic", "phylum", "class")] #column to pander
top_sensitive_microplastic$means_CTRL <- paste0(round(top_sensitive_microplastic$means_CTRL,2),"%")
top_sensitive_microplastic$means_microplastic <- paste0(round(top_sensitive_microplastic$means_microplastic,2),"%")
colnames(top_sensitive_microplastic)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_microplastic, caption="Top abundant ASVs sensitive to microplastic in root_sand")

```

\pagebreak

```{r root_sand: relative abundance taxa macroplastic, echo=F, message=F, warning=F}

#macroplastic

#sensitive ASVs from top 50
top_sensitive_macroplastic <- pivot_wider(summary_root_sand_macroplastic,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_macroplastic <- top_sensitive_macroplastic[1:top_abu,] #top abundant
top_sensitive_macroplastic <- top_sensitive_macroplastic[top_sensitive_macroplastic$sensitive_macroplastic =="sensitive",] #sensitive
top_sensitive_macroplastic <- top_sensitive_macroplastic[,colnames(top_sensitive_macroplastic) %in%
                                                         c("ASV","means_CTRL","means_macroplastic", "phylum", "class")] #column to pander
top_sensitive_macroplastic$means_CTRL <- paste0(round(top_sensitive_macroplastic$means_CTRL,2),"%")
top_sensitive_macroplastic$means_macroplastic <- paste0(round(top_sensitive_macroplastic$means_macroplastic,2),"%")
colnames(top_sensitive_macroplastic)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_macroplastic, caption="Top abundant ASVs sensitive to macroplastic in root_sand")

```

\pagebreak

```{r root_sand: relative abundance taxa starch, echo=F, message=F, warning=F}

#starch

#sensitive ASVs from top 50
top_sensitive_starch <- pivot_wider(summary_root_sand_starch,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_starch <- top_sensitive_starch[1:top_abu,] #top abundant
top_sensitive_starch <- top_sensitive_starch[top_sensitive_starch$sensitive_starch =="sensitive",] #sensitive
top_sensitive_starch <- top_sensitive_starch[,colnames(top_sensitive_starch) %in%
                                                         c("ASV","means_CTRL","means_starch", "phylum", "class")] #column to pander
top_sensitive_starch$means_CTRL <- paste0(round(top_sensitive_starch$means_CTRL,2),"%")
top_sensitive_starch$means_starch <- paste0(round(top_sensitive_starch$means_starch,2),"%")
colnames(top_sensitive_starch)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_starch, caption="Top abundant ASVs sensitive to starch in root_sand")

```

\pagebreak

### Sensitive Taxonomies

We plot how many ASVs are sensitive to multiple plastic treatments and show frequent phyla, classes and families which are sensitive to all three plastic treatments.

\vspace{5mm}

```{r root_sand: Venn plot, echo=F, message=F, warning=F}

ggvenn(list(microplastic=summary_root_sand_microplastic$ASV[summary_root_sand_microplastic$sensitive=="sensitive"],
            macroplastic=summary_root_sand_macroplastic$ASV[summary_root_sand_macroplastic$sensitive=="sensitive"],
            starch=summary_root_sand_starch$ASV[summary_root_sand_starch$sensitive=="sensitive"]),
  fill_color = c("darkolivegreen2", 'darkseagreen4', 'gold'),
  stroke_size = 0.5, set_name_size = 5, 
  #set_name_color = c("darkolivegreen2", 'darkseagreen4', 'gold'),
  show_percentage = F
  )+
  ggtitle("root_sand: Sensitive bacterial ASVs in root_sand")

```

\pagebreak

```{r root_sand: common frequent sensitive taxa, echo=F, message=F, warning=F}

#sensitive taxonomy frequent in microplastic, macroplastic and starch
sensitive_common_phylum <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV, family), names_from = "Plastic", values_from = c("phylum")) %>%
                             na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
colnames(sensitive_common_phylum)[2] <- "phylum"
sensitive_common_phylum <- as.data.frame(sort(table(sensitive_common_phylum$phylum),decreasing = T)) %>% filter(Freq >= 5) #filter by phylum with more than 4 sensitive ASVs
colnames(sensitive_common_phylum) <- c("phylum", "frequency")
pander(sensitive_common_phylum, caption="Frequent phyla sensitive to all plastic treatments  in root_sand")

#class
sensitive_common_class <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV, family), names_from = "Plastic", values_from = c("class")) %>%
                             na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
colnames(sensitive_common_class)[2] <- "class"
sensitive_common_class <- as.data.frame(sort(table(sensitive_common_class$class),decreasing = T)) %>% filter(Freq >= 5) #filter by class with more than 4 sensitive ASVs
colnames(sensitive_common_class) <- c("class", "frequency")
pander(sensitive_common_class, caption="Frequent classes sensitive to all plastic treatments in root_sand")

#family
sensitive_common_family <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV), names_from = "Plastic", values_from = c("family")) %>%
                             na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
colnames(sensitive_common_family)[2] <- "family"
sensitive_common_family <- as.data.frame(sort(table(sensitive_common_family$family),decreasing = T)) %>% filter(Freq >= 5) #filter by family with more than 4 sensitive ASVs
colnames(sensitive_common_family) <- c("family", "frequency")
pander(sensitive_common_family, caption="Frequent families sensitive to all plastic treatments in root_sand")

```

\pagebreak

### Phylogenetic tree

We plot a phylogenetic tree with all ASVs tested for sensitivity and label with the rings around the tree if they are sensitive to microplastic, macroplastic and starch.

\vspace{5mm}

```{r root_sand: subset sequences for tree, echo=F, message=F, warning=F}

#import seqs
seqs <- Biostrings::readDNAStringSet("../01_dadapipe/4_output/bacteria_ASV/ASV100/bacteria_SEQ100.fasta")
seqs <- as.data.frame(seqs); colnames(seqs) <- "seqs"

#subset for tree
tested_ASVs_root_sand <- unique(summ$ASV)
seqs_root_sand <- seqs[rownames(seqs) %in% tested_ASVs_root_sand,,drop=F]
seqs_root_sand <- data.frame(name=rownames(seqs_root_sand), seq=seqs_root_sand$seqs)
rownames(seqs_root_sand) <- seqs_root_sand$name


#export as fasta file
#dir.create("../03_build_tree/root_sand",recursive = T)
#ampir::df_to_faa(seqs_root_sand, "../03_build_tree/root_sand/sequences_root_sand.fasta")

########
#in bash
#cd ../03_build_tree/root_sand/sequences_root_sand
#mafft-linsi sequences_root_sand.fasta > sequences_root_sand_aligned.fasta #align sequences
#iqtree -s sequences_root_sand_aligned.fasta #build tree
########
```

```{r root_sand: show taxa tree, echo=F, message=F, warning=F}

#set to true at the beginning of script if files were computed
if(treefiles){

  #read tree
  tree_file <- "/Users/j.waechli/Desktop/Jan_Waelchli/01_Projects/01_Support/Robin/Microbiome_BE13/03_build_tree/root_sand/sequences_root_sand_aligned.fasta.treefile"
  tree <- read.tree(tree_file)
  
  #subset summary and sort them
  summary_taxa <- unique(summary[,c("ASV", "phylum","class","family")])
  summary_taxa <- summary_taxa[summary_taxa$ASV %in% tested_ASVs_root_sand,]
  rownames(summary_taxa) <- summary_taxa$ASV
  summary_taxa <- summary_taxa[tree[["tip.label"]],]
  
  
  #colorstring
  summary_taxa$labels <- NA
  summary_taxa$cols <- NA
  
  for(a in summary_taxa$ASV){
    summary_taxa$labels[summary_taxa$ASV == a] <- as.character(TAX$labels[TAX$ASV_ID == a][1])
    summary_taxa$cols[summary_taxa$ASV == a] <- as.character(TAX$cols_phyla[TAX$ASV_ID == a][1])
  }
  
  high_abundant_phyla <- as.character(unique(PHYSEQ_phyla_melt$labels_2))
  summary_taxa$labels_2 <- summary_taxa$labels
  summary_taxa$labels_2[!(summary_taxa$labels %in% high_abundant_phyla)] <- "Low abundant phyla"
  summary_taxa$cols2 <- summary_taxa$cols
  summary_taxa$cols2[summary_taxa$labels_2 == "Low abundant phyla"] <- "lightgrey"
  
  #extend taxa string to the length of tree edges (needs for each edge a value but uses only the number of tips)
  phylum <- c(summary_taxa$labels_2,rep(NA,(length(tree[["edge.length"]])-length(summary_taxa$labels_2))+1))
  cols2 <- summary_taxa$cols2
  names(cols2) <- summary_taxa$labels_2
  cols2_long <- c(cols2,rep("white",(length(tree[["edge.length"]])-length(cols2))+1))
  
  #plot genetic tree with phyla
  p1 <- ggtree(tree, layout = "fan", branch.length = "none")+
    geom_tippoint(aes(color=phylum), size=2)+
    scale_color_manual(values=cols2_long)+
    new_scale_color()
  
  
  #for saving
  phylotree_root_sand <- list(plot=p1, phylum=phylum, colstring=cols2_long)
  
  
  #plot microplastic ring
  df <- summary[summary$Plastic %in% c("CTRL", "microplastic") & summary$ASV %in% rownames(seqs_root_sand),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_microplastic) / (df$means_CTRL + df$means_microplastic)) *100
  df$abu_trt_microplastic <- factor(df$abu_trt_microplastic, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p2 <- p1 + geom_rect(aes(xmin = 45, xmax = 67, ymin = 1, ymax = nrow(df)), fill="darkolivegreen2")+ #background color
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_microplastic),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list()) +
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  #plot macroplastic ring
  df <- summary[summary$Plastic %in% c("CTRL", "macroplastic") & summary$ASV %in% rownames(seqs_root_sand),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_macroplastic) / (df$means_CTRL + df$means_macroplastic)) *100
  df$abu_trt_macroplastic <- factor(df$abu_trt_macroplastic, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p3 <- p2 + geom_rect(aes(xmin = 67, xmax = 88, ymin = 1, ymax = nrow(df)), fill="darkseagreen4")+
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_macroplastic),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list())+
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  
  #plot starch ring
  df <- summary[summary$Plastic %in% c("CTRL", "starch") & summary$ASV %in% rownames(seqs_root_sand),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_starch) / (df$means_CTRL + df$means_starch)) *100
  df$abu_trt_starch <- factor(df$abu_trt_starch, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p4 <- p3 + geom_rect(aes(xmin = 88, xmax = 110, ymin = 1, ymax = nrow(df)), fill="gold")+
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_starch),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list())+
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  #pseudo plot for legend
  library(ggnewscale)
  df <- data.frame(Phylum=names(cols2), Plastic=c(rep("microplastic",length(cols2)-2),"macroplastic","starch"),
                   Sensitivity=c(rep("higher in CTRL",length(cols2)-2),"unchanged","lower in CTRL"))
  df$Sensitivity <- factor(df$Sensitivity, levels=c("higher in CTRL", "unchanged","lower in CTRL"))
  legend_plot <- ggplot(df,aes(1,1))+
                    geom_bar(stat="identity", aes(fill=Plastic))+
                    geom_point(pch=15, size=6, aes(col=Phylum))+
                    scale_color_manual(values=cols2)+
                    scale_fill_manual(values=level_cols_Plastic[c(1,2,4)], name = "Plastic")+
                    theme_bw()+
                    new_scale_color() +
                    geom_point(pch=15, size=6, aes(col=Sensitivity))+
                    scale_color_manual(values=c("tomato", "grey80", "darkcyan"))
                    #guides(size = guide_legend(order = 1), color = guide_legend(order = 2))
  
  #add plot and legend
  phyla_plot <- plot_grid(p4 + theme(legend.position="none"), 
                          get_legend(legend_plot), nrow=1,rel_widths=c(1, .3))
  phyla_plot
  
}

```

\pagebreak

## Relative abundance in root soil

We check for each ASV if it is sensitive or not to the different plastics. We show how many sensitive ASVs we have found for each plastic-treatment. Then, we plot the top 50 most abundant ASVs and label significant differences in red. Down below we show the phyla and class of those top abundant sensitive ASVs. 

```{r root_soil: compute DAA, echo=F, message=F, warning=F}

#microplastic

#subset to microplastic
DESIGN_root_soil_microplastic <-  DESIGN_root_soil[(DESIGN_root_soil$Plastic %in% c("CTRL", "microplastic") & DESIGN_root_soil$Plant==T),]
DAT_root_soil_microplastic <- DAT_root_soil[colnames(DAT_root_soil) %in% DESIGN_root_soil_microplastic$AccessArray_barcode] #for DAA analysis

#preprocess
temp <- preprocess(DESIGN_root_soil_microplastic, DAT_root_soil_microplastic, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_root_soil_microplastic <- temp[[1]]; rm(temp)
sens_ASVs_root_soil_microplastic <- summ_root_soil_microplastic$ASV[summ_root_soil_microplastic$score>2]


#macroplastic

#subset to macroplastic
DESIGN_root_soil_macroplastic <-  DESIGN_root_soil[(DESIGN_root_soil$Plastic %in% c("CTRL", "macroplastic") & DESIGN_root_soil$Plant==T),]
DAT_root_soil_macroplastic <- DAT_root_soil[colnames(DAT_root_soil) %in% DESIGN_root_soil_macroplastic$AccessArray_barcode] #forr DAA analysis

#preprocess
temp <- preprocess(DESIGN_root_soil_macroplastic, DAT_root_soil_macroplastic, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_root_soil_macroplastic <- temp[[1]]; rm(temp)
sens_ASVs_root_soil_macroplastic <- summ_root_soil_macroplastic$ASV[summ_root_soil_macroplastic$score>2]


#starch

#subset to starch
DESIGN_root_soil_starch <-  DESIGN_root_soil[(DESIGN_root_soil$Plastic %in% c("CTRL", "starch") & DESIGN_root_soil$Plant==T),]
DAT_root_soil_starch <- DAT_root_soil[colnames(DAT_root_soil) %in% DESIGN_root_soil_starch$AccessArray_barcode] #forr DAA analysis

#preprocess
temp <- preprocess(DESIGN_root_soil_starch, DAT_root_soil_starch, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_root_soil_starch <- temp[[1]]; rm(temp)
sens_ASVs_root_soil_starch <- summ_root_soil_starch$ASV[summ_root_soil_starch$score>2]

```


```{r root_soil: summarise DAA, echo=F, message=F, warning=F}

#relative abundance
DAT_root_soil_rare <- as.data.frame(DAT_root_soil_rare)
DAT_root_soil_rare <- (DAT_root_soil_rare/colSums(DAT_root_soil_rare)) * 100
DAT_root_soil_rare <- as.data.frame(DAT_root_soil_rare[ rowSums(DAT_root_soil_rare) > 0,])

#means & standard error
means <- aggregate(t(DAT_root_soil_rare), list(DESIGN_root_soil$Plastic), mean) %>%
  column_to_rownames("Group.1") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("ASV")

se <- aggregate(t(DAT_root_soil_rare), list(DESIGN_root_soil$Plastic), se) %>%
  column_to_rownames("Group.1") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("ASV")

#sort by abundance of CTRL samples
order <- order(-means$CTRL)
means <- means[order,]
se <- se[order,]

summary <- pivot_longer(means, !(ASV), names_to="Plastic", values_to="means")
summary$se <- pivot_longer(se, !(ASV), names_to="Plastic", values_to="se") %>% pull(se)
summary$Plastic <- factor(summary$Plastic, levels=c("CTRL", "microplastic", "macroplastic", "starch"))
summary$ASV <- factor(summary$ASV, levels=unique(summary$ASV)) #keep the sorted order of ASVs

#add score
#score
summ_root_soil_microplastic$plastic <- "microplastic"
summ_root_soil_macroplastic$plastic <- "macroplastic"
summ_root_soil_starch$plastic <- "starch"
summ <- rbind(summ_root_soil_microplastic, summ_root_soil_macroplastic, summ_root_soil_starch)

#get score
summary$score <- 0
for(p in unique(summary$Plastic)){
  for(a in unique(summary$ASV)){
    score <- summ %>% filter(plastic == p) %>% filter(ASV == a) %>% pull(score)
    if(length(score)==1){summary$score[(summary$Plastic==p & summary$ASV==a)] <- score}
  }}

#add if different abundant (score > 2)
summary$sensitive <- ifelse(summary$score > 2, "sensitive", "not_sensitive")
summary$sensitive <- as.factor(summary$sensitive)

#add abundace in trt compared to ctrl
summary$abu_trt <- "unchanged"
for(p in unique(summary$Plastic)){
  for(a in unique(summary$ASV)){
    mean_ctrl <- summary$means[(summary$ASV==a) & (summary$Plastic=="CTRL")]
    mean_trt <- summary$means[(summary$ASV==a) & (summary$Plastic==p)]
    abu_sens <- summary$sensitive[(summary$ASV==a) & (summary$Plastic==p)]
    if(abu_sens == "sensitive"){
      trt_dir <- ifelse(mean_ctrl > mean_trt, "higher in CTRL", "lower in CTRL")
      summary$abu_trt[(summary$ASV==a) & (summary$Plastic==p)] <- trt_dir
      }
  }
}

#add phyla and class
summary$phylum <- NA
summary$class <- NA
summary$order <- NA
summary$family <- NA
summary$genus <- NA
for(a in unique(summary$ASV)){
  if(a %in% TAX$ASV_ID){
    summary$phylum[summary$ASV == a] <- as.character(TAX$phylum[TAX$ASV_ID == a])
    summary$class[summary$ASV == a] <- as.character(TAX$class[TAX$ASV_ID == a])
    summary$order[summary$ASV == a] <- as.character(TAX$order[TAX$ASV_ID == a])
    summary$family[summary$ASV == a] <- as.character(TAX$family[TAX$ASV_ID == a])
    summary$genus[summary$ASV == a] <- as.character(TAX$genus[TAX$ASV_ID == a])
  }
  else{
    summary$phylum[summary$ASV == a] <- "unassigned"
    summary$class[summary$ASV == a] <- "unassigned"
    summary$order[summary$ASV == a] <- "unassigned"
    summary$family[summary$ASV == a] <- "unassigned"
    summary$genus[summary$ASV == a] <- "unassigned"
  }
}

summary_root_soil <- summary

```

### Different abundant ASVs

We show how many ASVs has been changed due to the plastic treatments how much of the relative abundance belongs to those sensitive ASVs.

```{r root_soil: show DAA microplastic, echo=F, message=F, warning=F}

#pander how many ASVs are sensitive

#microplastic
summary_root_soil_microplastic <- summary[(summary$Plastic %in% c("CTRL", "microplastic")) & (summary$ASV %in% summ_root_soil_microplastic$ASV),]
changed_abu_root_soil_microplastic <- sum(summary_root_soil_microplastic$means[summary_root_soil_microplastic$Plastic=="microplastic" &
                                      summary_root_soil_microplastic$sensitive=="sensitive"])
changed_abu_root_soil_microplastic <- paste0(round(changed_abu_root_soil_microplastic,2), "%")
sens_root_soil_microplastic <- table(summary_root_soil_microplastic$abu_trt[summary_root_soil_microplastic$Plastic == "microplastic"])
pander(sens_root_soil_microplastic[c(2,3,1)], caption = "microplastic effect in root_soil")

```

`r changed_abu_root_soil_microplastic` of the bacterial community in root was changed in abundance due to microplastic.

```{r root_soil: show DAA macroplastic, echo=F, message=F, warning=F}

#macroplastic
summary_root_soil_macroplastic <- summary[(summary$Plastic %in% c("CTRL", "macroplastic")) & (summary$ASV %in% summ_root_soil_macroplastic$ASV),]
changed_abu_root_soil_macroplastic <- sum(summary_root_soil_macroplastic$means[summary_root_soil_macroplastic$Plastic=="macroplastic" &
                                      summary_root_soil_macroplastic$sensitive=="sensitive"])
changed_abu_root_soil_macroplastic <- paste0(round(changed_abu_root_soil_macroplastic,2), "%")
sens_root_soil_macroplastic <- table(summary_root_soil_macroplastic$abu_trt[summary_root_soil_macroplastic$Plastic == "macroplastic"])
pander(sens_root_soil_macroplastic[c(2,3,1)], caption = "macroplastic effect in root_soil")

```

`r changed_abu_root_soil_macroplastic` of the bacterial community in root was changed in abundance due to macroplastic.

```{r root_soil: show DAA starch, echo=F, message=F, warning=F}

#starch
summary_root_soil_starch <- summary[(summary$Plastic %in% c("CTRL", "starch")) & (summary$ASV %in% summ_root_soil_starch$ASV),]
changed_abu_root_soil_starch <- sum(summary_root_soil_starch$means[summary_root_soil_starch$Plastic=="starch" &
                                      summary_root_soil_starch$sensitive=="sensitive"])
changed_abu_root_soil_starch <- paste0(round(changed_abu_root_soil_starch,2), "%")
sens_root_soil_starch <- table(summary_root_soil_starch$abu_trt[summary_root_soil_starch$Plastic == "starch"])
pander(sens_root_soil_starch[c(2,3,1)], caption = "starch effect in root_soil")

```

`r changed_abu_root_soil_starch` of the bacterial community in root_soil was changed in abundance due to starch.

\pagebreak

### Relative abundance plot

We plot the 50 top abundant ASVs (in controls) and label plastic sensitive ASVs in red.

\vspace{5mm}

```{r root_soil: relative abundance plot, echo=F, message=F, warning=F}

#plot the 50 most abundant ASVs

#set threshold for top abundant
top_abu <- 50

#plot
ggplot(summary[1:(top_abu*4),], aes(ASV, means, fill=Plastic, col=sensitive))+
  geom_bar(stat="identity", position="dodge", width = 0.8, size=0.2)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, size=6))+
  scale_color_manual(values=c(NA, "red"))+
  scale_fill_manual(values=level_cols_Plastic)+
  geom_errorbar(aes(ymin=means, ymax=means+se/2), width=.3, position=position_dodge(.8))
  #geom_text(aes(label = score, y=-0.2), position = position_dodge(.8), size=1)

```

\pagebreak

### Taxonomy of abundant ASVs

We show the taxonomy of top abundant ASVs which are plastic sensitive.

```{r root_soil: relative abundance taxa microplastic, echo=F, message=F, warning=F}

#microplastic

#sensitive ASVs from top 50
top_sensitive_microplastic <- pivot_wider(summary_root_soil_microplastic,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_microplastic <- top_sensitive_microplastic[1:top_abu,] #top abundant
top_sensitive_microplastic <- top_sensitive_microplastic[top_sensitive_microplastic$sensitive_microplastic =="sensitive",] #sensitive
top_sensitive_microplastic <- top_sensitive_microplastic[,colnames(top_sensitive_microplastic) %in%
                                                         c("ASV","means_CTRL","means_microplastic", "phylum", "class")] #column to pander
top_sensitive_microplastic$means_CTRL <- paste0(round(top_sensitive_microplastic$means_CTRL,2),"%")
top_sensitive_microplastic$means_microplastic <- paste0(round(top_sensitive_microplastic$means_microplastic,2),"%")
colnames(top_sensitive_microplastic)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_microplastic, caption="Top abundant ASVs sensitive to microplastic in root_soil")

```

\pagebreak

```{r root_soil: relative abundance taxa macroplastic, echo=F, message=F, warning=F}

#macroplastic

#sensitive ASVs from top 50
top_sensitive_macroplastic <- pivot_wider(summary_root_soil_macroplastic,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_macroplastic <- top_sensitive_macroplastic[1:top_abu,] #top abundant
top_sensitive_macroplastic <- top_sensitive_macroplastic[top_sensitive_macroplastic$sensitive_macroplastic =="sensitive",] #sensitive
top_sensitive_macroplastic <- top_sensitive_macroplastic[,colnames(top_sensitive_macroplastic) %in%
                                                         c("ASV","means_CTRL","means_macroplastic", "phylum", "class")] #column to pander
top_sensitive_macroplastic$means_CTRL <- paste0(round(top_sensitive_macroplastic$means_CTRL,2),"%")
top_sensitive_macroplastic$means_macroplastic <- paste0(round(top_sensitive_macroplastic$means_macroplastic,2),"%")
colnames(top_sensitive_macroplastic)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_macroplastic, caption="Top abundant ASVs sensitive to macroplastic in root_soil")

```

\pagebreak

```{r root_soil: relative abundance taxa starch, echo=F, message=F, warning=F}

#starch

#sensitive ASVs from top 50
top_sensitive_starch <- pivot_wider(summary_root_soil_starch,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_starch <- top_sensitive_starch[1:top_abu,] #top abundant
top_sensitive_starch <- top_sensitive_starch[top_sensitive_starch$sensitive_starch =="sensitive",] #sensitive
top_sensitive_starch <- top_sensitive_starch[,colnames(top_sensitive_starch) %in%
                                                         c("ASV","means_CTRL","means_starch", "phylum", "class")] #column to pander
top_sensitive_starch$means_CTRL <- paste0(round(top_sensitive_starch$means_CTRL,2),"%")
top_sensitive_starch$means_starch <- paste0(round(top_sensitive_starch$means_starch,2),"%")
colnames(top_sensitive_starch)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_starch, caption="Top abundant ASVs sensitive to starch in root_soil")

```

\pagebreak

### Sensitive Taxonomies

We plot how many ASVs are sensitive to multiple plastic treatments and show frequent phyla, classes and families which are sensitive to all three plastic treatments.

\vspace{5mm}

```{r root_soil: Venn plot, echo=F, message=F, warning=F}

ggvenn(list(microplastic=summary_root_soil_microplastic$ASV[summary_root_soil_microplastic$sensitive=="sensitive"],
            macroplastic=summary_root_soil_macroplastic$ASV[summary_root_soil_macroplastic$sensitive=="sensitive"],
            starch=summary_root_soil_starch$ASV[summary_root_soil_starch$sensitive=="sensitive"]),
  fill_color = c("darkolivegreen2", 'darkseagreen4', 'gold'),
  stroke_size = 0.5, set_name_size = 5, 
  #set_name_color = c("darkolivegreen2", 'darkseagreen4', 'gold'),
  show_percentage = F
  )+
  ggtitle("root_soil: Sensitive bacterial ASVs in root_soil")

```

\pagebreak

```{r root_soil: common frequent sensitive taxa, echo=F, message=F, warning=F}

# #sensitive taxonomy frequent in microplastic, macroplastic and starch
# sensitive_common_phylum <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV, family), names_from = "Plastic", values_from = c("phylum")) %>%
#                              na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
# colnames(sensitive_common_phylum)[2] <- "phylum"
# sensitive_common_phylum <- as.data.frame(sort(table(sensitive_common_phylum$phylum),decreasing = T)) %>% filter(Freq >= 5) #filter by phylum with more than 4 sensitive ASVs
# colnames(sensitive_common_phylum) <- c("phylum", "frequency")
# pander(sensitive_common_phylum, caption="Frequent phyla sensitive to all plastic treatments  in root_soil")
# 
# #class
# sensitive_common_class <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV, family), names_from = "Plastic", values_from = c("class")) %>%
#                              na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
# colnames(sensitive_common_class)[2] <- "class"
# sensitive_common_class <- as.data.frame(sort(table(sensitive_common_class$class),decreasing = T)) %>% filter(Freq >= 5) #filter by class with more than 4 sensitive ASVs
# colnames(sensitive_common_class) <- c("class", "frequency")
# pander(sensitive_common_class, caption="Frequent classes sensitive to all plastic treatments in root_soil")
# 
# #family
# sensitive_common_family <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV), names_from = "Plastic", values_from = c("family")) %>%
#                              na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
# colnames(sensitive_common_family)[2] <- "family"
# sensitive_common_family <- as.data.frame(sort(table(sensitive_common_family$family),decreasing = T)) %>% filter(Freq >= 5) #filter by family with more than 4 sensitive ASVs
# colnames(sensitive_common_family) <- c("family", "frequency")
# pander(sensitive_common_family, caption="Frequent families sensitive to all plastic treatments in root_soil")

```

\pagebreak

### Phylogenetic tree

We plot a phylogenetic tree with all ASVs tested for sensitivity and label with the rings around the tree if they are sensitive to microplastic, macroplastic and starch.

\vspace{5mm}

```{r root_soil: subset sequences for tree, echo=F, message=F, warning=F}

#import seqs
seqs <- Biostrings::readDNAStringSet("../01_dadapipe/4_output/bacteria_ASV/ASV100/bacteria_SEQ100.fasta")
seqs <- as.data.frame(seqs); colnames(seqs) <- "seqs"

#subset for tree
tested_ASVs_root_soil <- unique(summ$ASV)
seqs_root_soil <- seqs[rownames(seqs) %in% tested_ASVs_root_soil,,drop=F]
seqs_root_soil <- data.frame(name=rownames(seqs_root_soil), seq=seqs_root_soil$seqs)
rownames(seqs_root_soil) <- seqs_root_soil$name


#export as fasta file
#dir.create("../03_build_tree/root_soil",recursive = T)
#ampir::df_to_faa(seqs_root_soil, "../03_build_tree/root_soil/sequences_root_soil.fasta")

########
#in bash
#cd ../03_build_tree/root_soil/sequences_root_soil
#mafft-linsi sequences_root_soil.fasta > sequences_root_soil_aligned.fasta #align sequences
#iqtree -s sequences_root_soil_aligned.fasta #build tree
########
```

```{r root_soil: show taxa tree, echo=F, message=F, warning=F}

#set to true at the beginning of script if files were computed
if(treefiles){

  #read tree
  tree_file <- "/Users/j.waechli/Desktop/Jan_Waelchli/01_Projects/01_Support/Robin/Microbiome_BE13/03_build_tree/root_soil/sequences_root_soil_aligned.fasta.treefile"
  tree <- read.tree(tree_file)
  
  #subset summary and sort them
  summary_taxa <- unique(summary[,c("ASV", "phylum","class","family")])
  summary_taxa <- summary_taxa[summary_taxa$ASV %in% tested_ASVs_root_soil,]
  rownames(summary_taxa) <- summary_taxa$ASV
  summary_taxa <- summary_taxa[tree[["tip.label"]],]
  
  
  #colorstring
  summary_taxa$labels <- NA
  summary_taxa$cols <- NA
  
  for(a in summary_taxa$ASV){
    summary_taxa$labels[summary_taxa$ASV == a] <- as.character(TAX$labels[TAX$ASV_ID == a][1])
    summary_taxa$cols[summary_taxa$ASV == a] <- as.character(TAX$cols_phyla[TAX$ASV_ID == a][1])
  }
  
  high_abundant_phyla <- as.character(unique(PHYSEQ_phyla_melt$labels_2))
  summary_taxa$labels_2 <- summary_taxa$labels
  summary_taxa$labels_2[!(summary_taxa$labels %in% high_abundant_phyla)] <- "Low abundant phyla"
  summary_taxa$cols2 <- summary_taxa$cols
  summary_taxa$cols2[summary_taxa$labels_2 == "Low abundant phyla"] <- "lightgrey"
  
  #extend taxa string to the length of tree edges (needs for each edge a value but uses only the number of tips)
  phylum <- c(summary_taxa$labels_2,rep(NA,(length(tree[["edge.length"]])-length(summary_taxa$labels_2))+1))
  cols2 <- summary_taxa$cols2
  names(cols2) <- summary_taxa$labels_2
  cols2_long <- c(cols2,rep("white",(length(tree[["edge.length"]])-length(cols2))+1))
  
  #plot genetic tree with phyla
  p1 <- ggtree(tree, layout = "fan", branch.length = "none")+
    geom_tippoint(aes(color=phylum), size=2)+
    scale_color_manual(values=cols2_long)+
    new_scale_color()
  
  
  #for saving
  phylotree_root_soil <- list(plot=p1, phylum=phylum, colstring=cols2_long)
  
  
  #plot microplastic ring
  df <- summary[summary$Plastic %in% c("CTRL", "microplastic") & summary$ASV %in% rownames(seqs_root_soil),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_microplastic) / (df$means_CTRL + df$means_microplastic)) *100
  df$abu_trt_microplastic <- factor(df$abu_trt_microplastic, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p2 <- p1 + geom_rect(aes(xmin = 42, xmax = 60, ymin = 1, ymax = nrow(df)), fill="darkolivegreen2")+ #background color
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_microplastic),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list()) +
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
              #scale_fill_manual(values=c("grey80"))
  
  #plot macroplastic ring
  df <- summary[summary$Plastic %in% c("CTRL", "macroplastic") & summary$ASV %in% rownames(seqs_root_soil),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_macroplastic) / (df$means_CTRL + df$means_macroplastic)) *100
  df$abu_trt_macroplastic <- factor(df$abu_trt_macroplastic, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p3 <- p2 + geom_rect(aes(xmin = 60, xmax = 83, ymin = 1, ymax = nrow(df)), fill="darkseagreen4")+
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_macroplastic),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list())+
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  
  #plot starch ring
  df <- summary[summary$Plastic %in% c("CTRL", "starch") & summary$ASV %in% rownames(seqs_root_soil),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_starch) / (df$means_CTRL + df$means_starch)) *100
  df$abu_trt_starch <- factor(df$abu_trt_starch, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p4 <- p3 + geom_rect(aes(xmin = 83, xmax = 107, ymin = 1, ymax = nrow(df)), fill="gold")+
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_starch),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list())+
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  #pseudo plot for legend
  library(ggnewscale)
  df <- data.frame(Phylum=names(cols2), Plastic=c(rep("microplastic",length(cols2)-2),"macroplastic","starch"),
                   Sensitivity=c(rep("higher in CTRL",length(cols2)-2),"unchanged","lower in CTRL"))
  df$Sensitivity <- factor(df$Sensitivity, levels=c("higher in CTRL", "unchanged","lower in CTRL"))
  legend_plot <- ggplot(df,aes(1,1))+
                    geom_bar(stat="identity", aes(fill=Plastic))+
                    geom_point(pch=15, size=6, aes(col=Phylum))+
                    scale_color_manual(values=cols2)+
                    scale_fill_manual(values=level_cols_Plastic[c(1,2,4)], name = "Plastic")+
                    theme_bw()+
                    new_scale_color() +
                    geom_point(pch=15, size=6, aes(col=Sensitivity))+
                    scale_color_manual(values=c("tomato", "grey80", "darkcyan"))
                    #guides(size = guide_legend(order = 1), color = guide_legend(order = 2))
  
  #add plot and legend
  phyla_plot <- plot_grid(p4 + theme(legend.position="none"), 
                          get_legend(legend_plot), nrow=1,rel_widths=c(1, .3))
  phyla_plot
  
}

```

\pagebreak


## Relative abundance in soil

We check for each ASV if it is sensitive or not to the different plastics. We show how many sensitive ASVs we have found for each plastic-treatment. Then, we plot the top 50 most abundant ASVs and label significant differences in red. Down below we show the phyla and class of those top abundant sensitive ASVs. 

```{r soil: compute DAA, echo=F, message=F, warning=F}

#microplastic

#subset to microplastic
DESIGN_soil_microplastic <-  DESIGN_soil[(DESIGN_soil$Plastic %in% c("CTRL", "microplastic") & DESIGN_soil$Plant==T),]
DAT_soil_microplastic <- DAT_soil[colnames(DAT_soil) %in% DESIGN_soil_microplastic$AccessArray_barcode] #for DAA analysis

#preprocess
temp <- preprocess(DESIGN_soil_microplastic, DAT_soil_microplastic, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_soil_microplastic <- temp[[1]]; rm(temp)
sens_ASVs_soil_microplastic <- summ_soil_microplastic$ASV[summ_soil_microplastic$score>2]


#macroplastic

#subset to macroplastic
DESIGN_soil_macroplastic <-  DESIGN_soil[(DESIGN_soil$Plastic %in% c("CTRL", "macroplastic") & DESIGN_soil$Plant==T),]
DAT_soil_macroplastic <- DAT_soil[colnames(DAT_soil) %in% DESIGN_soil_macroplastic$AccessArray_barcode] #forr DAA analysis

#preprocess
temp <- preprocess(DESIGN_soil_macroplastic, DAT_soil_macroplastic, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_soil_macroplastic <- temp[[1]]; rm(temp)
sens_ASVs_soil_macroplastic <- summ_soil_macroplastic$ASV[summ_soil_macroplastic$score>2]


#starch

#subset to starch
DESIGN_soil_starch <-  DESIGN_soil[(DESIGN_soil$Plastic %in% c("CTRL", "starch") & DESIGN_soil$Plant==T),]
DAT_soil_starch <- DAT_soil[colnames(DAT_soil) %in% DESIGN_soil_starch$AccessArray_barcode] #forr DAA analysis

#preprocess
temp <- preprocess(DESIGN_soil_starch, DAT_soil_starch, "AccessArray_barcode")
feature_table <- temp[[1]]; meta_data <- temp[[2]]; rm(temp)

#search differntial abundance
temp <- run_all_DAA(feature_table, meta_data, group = "Plastic", 
                    DAT_sample_names =  "AccessArray_barcode", reference = "Plastic,CTRL")

summ_soil_starch <- temp[[1]]; rm(temp)
sens_ASVs_soil_starch <- summ_soil_starch$ASV[summ_soil_starch$score>2]

```


```{r soil: summarise DAA, echo=F, message=F, warning=F}

#relative abundance
DAT_soil_rare <- as.data.frame(DAT_soil_rare)
DAT_soil_rare <- (DAT_soil_rare/colSums(DAT_soil_rare)) * 100
DAT_soil_rare <- as.data.frame(DAT_soil_rare[ rowSums(DAT_soil_rare) > 0,])

#means & standard error
means <- aggregate(t(DAT_soil_rare), list(DESIGN_soil$Plastic), mean) %>%
  column_to_rownames("Group.1") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("ASV")

se <- aggregate(t(DAT_soil_rare), list(DESIGN_soil$Plastic), se) %>%
  column_to_rownames("Group.1") %>%
  t() %>% as.data.frame() %>%
  rownames_to_column("ASV")

#sort by abundance of CTRL samples
order <- order(-means$CTRL)
means <- means[order,]
se <- se[order,]

summary <- pivot_longer(means, !(ASV), names_to="Plastic", values_to="means")
summary$se <- pivot_longer(se, !(ASV), names_to="Plastic", values_to="se") %>% pull(se)
summary$Plastic <- factor(summary$Plastic, levels=c("CTRL", "microplastic", "macroplastic", "starch"))
summary$ASV <- factor(summary$ASV, levels=unique(summary$ASV)) #keep the sorted order of ASVs

#add score
#score
summ_soil_microplastic$plastic <- "microplastic"
summ_soil_macroplastic$plastic <- "macroplastic"
summ_soil_starch$plastic <- "starch"
summ <- rbind(summ_soil_microplastic, summ_soil_macroplastic, summ_soil_starch)

#get score
summary$score <- 0
for(p in unique(summary$Plastic)){
  for(a in unique(summary$ASV)){
    score <- summ %>% filter(plastic == p) %>% filter(ASV == a) %>% pull(score)
    if(length(score)==1){summary$score[(summary$Plastic==p & summary$ASV==a)] <- score}
  }}

#add if different abundant (score > 2)
summary$sensitive <- ifelse(summary$score > 2, "sensitive", "not_sensitive")
summary$sensitive <- as.factor(summary$sensitive)

#add abundace in trt compared to ctrl
summary$abu_trt <- "unchanged"
for(p in unique(summary$Plastic)){
  for(a in unique(summary$ASV)){
    mean_ctrl <- summary$means[(summary$ASV==a) & (summary$Plastic=="CTRL")]
    mean_trt <- summary$means[(summary$ASV==a) & (summary$Plastic==p)]
    abu_sens <- summary$sensitive[(summary$ASV==a) & (summary$Plastic==p)]
    if(abu_sens == "sensitive"){
      trt_dir <- ifelse(mean_ctrl > mean_trt, "higher in CTRL", "lower in CTRL")
      summary$abu_trt[(summary$ASV==a) & (summary$Plastic==p)] <- trt_dir
      }
  }
}

#add phyla and class
summary$phylum <- NA
summary$class <- NA
summary$order <- NA
summary$family <- NA
summary$genus <- NA
for(a in unique(summary$ASV)){
  if(a %in% TAX$ASV_ID){
    summary$phylum[summary$ASV == a] <- as.character(TAX$phylum[TAX$ASV_ID == a])
    summary$class[summary$ASV == a] <- as.character(TAX$class[TAX$ASV_ID == a])
    summary$order[summary$ASV == a] <- as.character(TAX$order[TAX$ASV_ID == a])
    summary$family[summary$ASV == a] <- as.character(TAX$family[TAX$ASV_ID == a])
    summary$genus[summary$ASV == a] <- as.character(TAX$genus[TAX$ASV_ID == a])
  }
  else{
    summary$phylum[summary$ASV == a] <- "unassigned"
    summary$class[summary$ASV == a] <- "unassigned"
    summary$order[summary$ASV == a] <- "unassigned"
    summary$family[summary$ASV == a] <- "unassigned"
    summary$genus[summary$ASV == a] <- "unassigned"
  }
}

summary_soil <- summary

```

### Different abundant ASVs

We show how many ASVs has been changed due to the plastic treatments how much of the relative abundance belongs to those sensitive ASVs.

```{r soil: show DAA microplastic, echo=F, message=F, warning=F}

#pander how many ASVs are sensitive

#microplastic
summary_soil_microplastic <- summary[(summary$Plastic %in% c("CTRL", "microplastic")) & (summary$ASV %in% summ_soil_microplastic$ASV),]
changed_abu_soil_microplastic <- sum(summary_soil_microplastic$means[summary_soil_microplastic$Plastic=="microplastic" &
                                      summary_soil_microplastic$sensitive=="sensitive"])
changed_abu_soil_microplastic <- paste0(round(changed_abu_soil_microplastic,2), "%")
sens_soil_microplastic <- table(summary_soil_microplastic$abu_trt[summary_soil_microplastic$Plastic == "microplastic"])
pander(sens_soil_microplastic[c(2,3,1)], caption = "microplastic effect in soil")

```

`r changed_abu_soil_microplastic` of the bacterial community in soil was changed in abundance due to microplastic.

```{r soil: show DAA macroplastic, echo=F, message=F, warning=F}

#macroplastic
summary_soil_macroplastic <- summary[(summary$Plastic %in% c("CTRL", "macroplastic")) & (summary$ASV %in% summ_soil_macroplastic$ASV),]
changed_abu_soil_macroplastic <- sum(summary_soil_macroplastic$means[summary_soil_macroplastic$Plastic=="macroplastic" &
                                      summary_soil_macroplastic$sensitive=="sensitive"])
changed_abu_soil_macroplastic <- paste0(round(changed_abu_soil_macroplastic,2), "%")
sens_soil_macroplastic <- table(summary_soil_macroplastic$abu_trt[summary_soil_macroplastic$Plastic == "macroplastic"])
pander(sens_soil_macroplastic[c(2,3,1)], caption = "macroplastic effect in soil")

```

`r changed_abu_soil_macroplastic` of the bacterial community in soil was changed in abundance due to macroplastic.

```{r soil: show DAA starch, echo=F, message=F, warning=F}

#starch
summary_soil_starch <- summary[(summary$Plastic %in% c("CTRL", "starch")) & (summary$ASV %in% summ_soil_starch$ASV),]
changed_abu_soil_starch <- sum(summary_soil_starch$means[summary_soil_starch$Plastic=="starch" &
                                      summary_soil_starch$sensitive=="sensitive"])
changed_abu_soil_starch <- paste0(round(changed_abu_soil_starch,2), "%")
sens_soil_starch <- table(summary_soil_starch$abu_trt[summary_soil_starch$Plastic == "starch"])
pander(sens_soil_starch[c(2,3,1)], caption = "starch effect in soil")

```

`r changed_abu_soil_starch` of the bacterial community in soil was changed in abundance due to starch.

\pagebreak

### Relative abundance plot

We plot the 50 top abundant ASVs (in controls) and label plastic sensitive ASVs in red.

\vspace{5mm}

```{r soil: relative abundance plot, echo=F, message=F, warning=F}

#plot the 50 most abundant ASVs

#set threshold for top abundant
top_abu <- 50

#plot
ggplot(summary[1:(top_abu*4),], aes(ASV, means, fill=Plastic, col=sensitive))+
  geom_bar(stat="identity", position="dodge", width = 0.8, size=0.2)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, size=6))+
  scale_color_manual(values=c(NA, "red"))+
  scale_fill_manual(values=level_cols_Plastic)+
  geom_errorbar(aes(ymin=means, ymax=means+se/2), width=.3, position=position_dodge(.8))
  #geom_text(aes(label = score, y=-0.2), position = position_dodge(.8), size=1)

```

\pagebreak

### Taxonomy of abundant ASVs

We show the taxonomy of top abundant ASVs which are plastic sensitive.

```{r soil: relative abundance taxa microplastic, echo=F, message=F, warning=F}

#microplastic

#sensitive ASVs from top 50
top_sensitive_microplastic <- pivot_wider(summary_soil_microplastic,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_microplastic <- top_sensitive_microplastic[1:top_abu,] #top abundant
top_sensitive_microplastic <- top_sensitive_microplastic[top_sensitive_microplastic$sensitive_microplastic =="sensitive",] #sensitive
top_sensitive_microplastic <- top_sensitive_microplastic[,colnames(top_sensitive_microplastic) %in%
                                                         c("ASV","means_CTRL","means_microplastic", "phylum", "class")] #column to pander
top_sensitive_microplastic$means_CTRL <- paste0(round(top_sensitive_microplastic$means_CTRL,2),"%")
top_sensitive_microplastic$means_microplastic <- paste0(round(top_sensitive_microplastic$means_microplastic,2),"%")
colnames(top_sensitive_microplastic)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_microplastic, caption="Top abundant ASVs sensitive to microplastic in soil")

```

\pagebreak

```{r soil: relative abundance taxa macroplastic, echo=F, message=F, warning=F}

#macroplastic

#sensitive ASVs from top 50
top_sensitive_macroplastic <- pivot_wider(summary_soil_macroplastic,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_macroplastic <- top_sensitive_macroplastic[1:top_abu,] #top abundant
top_sensitive_macroplastic <- top_sensitive_macroplastic[top_sensitive_macroplastic$sensitive_macroplastic =="sensitive",] #sensitive
top_sensitive_macroplastic <- top_sensitive_macroplastic[,colnames(top_sensitive_macroplastic) %in%
                                                         c("ASV","means_CTRL","means_macroplastic", "phylum", "class")] #column to pander
top_sensitive_macroplastic$means_CTRL <- paste0(round(top_sensitive_macroplastic$means_CTRL,2),"%")
top_sensitive_macroplastic$means_macroplastic <- paste0(round(top_sensitive_macroplastic$means_macroplastic,2),"%")
colnames(top_sensitive_macroplastic)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_macroplastic, caption="Top abundant ASVs sensitive to macroplastic in soil")

```

\pagebreak

```{r soil: relative abundance taxa starch, echo=F, message=F, warning=F}

#starch

#sensitive ASVs from top 50
top_sensitive_starch <- pivot_wider(summary_soil_starch,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
top_sensitive_starch <- top_sensitive_starch[1:top_abu,] #top abundant
top_sensitive_starch <- top_sensitive_starch[top_sensitive_starch$sensitive_starch =="sensitive",] #sensitive
top_sensitive_starch <- top_sensitive_starch[,colnames(top_sensitive_starch) %in%
                                                         c("ASV","means_CTRL","means_starch", "phylum", "class")] #column to pander
top_sensitive_starch$means_CTRL <- paste0(round(top_sensitive_starch$means_CTRL,2),"%")
top_sensitive_starch$means_starch <- paste0(round(top_sensitive_starch$means_starch,2),"%")
colnames(top_sensitive_starch)[4:5] <- c("abu CTRL", "abu trt")
pander(top_sensitive_starch, caption="Top abundant ASVs sensitive to starch in soil")

```

\pagebreak

### Sensitive Taxonomies

We plot how many ASVs are sensitive to multiple plastic treatments and show frequent phyla, classes and families which are sensitive to all three plastic treatments.

\vspace{5mm}

```{r soil: Venn plot, echo=F, message=F, warning=F}

ggvenn(list(microplastic=summary_soil_microplastic$ASV[summary_soil_microplastic$sensitive=="sensitive"],
            macroplastic=summary_soil_macroplastic$ASV[summary_soil_macroplastic$sensitive=="sensitive"],
            starch=summary_soil_starch$ASV[summary_soil_starch$sensitive=="sensitive"]),
  fill_color = c("darkolivegreen2", 'darkseagreen4', 'gold'),
  stroke_size = 0.5, set_name_size = 5, 
  #set_name_color = c("darkolivegreen2", 'darkseagreen4', 'gold'),
  show_percentage = F
  )+
  ggtitle("soil: Sensitive bacterial ASVs in soil")

```

\pagebreak

```{r soil: common sensitive taxa, echo=F, message=F, warning=F}

#sensitive taxonomy in microplastic, macroplastic and starch

#phylum
sensitive_common_phylum <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV, family), names_from = "Plastic", values_from = c("phylum")) %>%
                             na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
colnames(sensitive_common_phylum)[2] <- "phylum"
sensitive_common_phylum <- as.data.frame(sort(table(sensitive_common_phylum$phylum),decreasing = T)) # %>% filter(Freq >= 5) #don't filter
colnames(sensitive_common_phylum) <- c("phylum", "frequency")
pander(sensitive_common_phylum, caption="Phyla sensitive to all plastic treatments  in soil")

#class
sensitive_common_class <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV, family), names_from = "Plastic", values_from = c("class")) %>%
                             na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
colnames(sensitive_common_class)[2] <- "class"
sensitive_common_class <- as.data.frame(sort(table(sensitive_common_class$class),decreasing = T)) # %>% filter(Freq >= 5)
colnames(sensitive_common_class) <- c("class", "frequency")
pander(sensitive_common_class, caption="Classes sensitive to all plastic treatments in soil")

#family
sensitive_common_family <- pivot_wider(summary[summary$sensitive=="sensitive",],id_cols = c(ASV), names_from = "Plastic", values_from = c("family")) %>%
                             na.omit() %>% dplyr::select("ASV", "microplastic") %>% as.data.frame()
colnames(sensitive_common_family)[2] <- "family"
sensitive_common_family <- as.data.frame(sort(table(sensitive_common_family$family),decreasing = T)) # %>% filter(Freq >= 5)
colnames(sensitive_common_family) <- c("family", "frequency")
pander(sensitive_common_family, caption="Families sensitive to all plastic treatments in soil")

```

\pagebreak

### Phylogenetic tree

We plot a phylogenetic tree with all ASVs tested for sensitivity and label with the rings around the tree if they are sensitive to microplastic, macroplastic and starch.

\vspace{5mm}

```{r soil: subset sequences for tree, echo=F, message=F, warning=F}

#import seqs
seqs <- Biostrings::readDNAStringSet("../01_dadapipe/4_output/bacteria_ASV/ASV100/bacteria_SEQ100.fasta")
seqs <- as.data.frame(seqs); colnames(seqs) <- "seqs"

#subset for tree
tested_ASVs_soil <- unique(summ$ASV)
seqs_soil <- seqs[rownames(seqs) %in% tested_ASVs_soil,,drop=F]
seqs_soil <- data.frame(name=rownames(seqs_soil), seq=seqs_soil$seqs)
rownames(seqs_soil) <- seqs_soil$name

#export as fasta file
#dir.create("../03_build_tree/soil",recursive = T)
#ampir::df_to_faa(seqs_soil, "../03_build_tree/soil/sequences_soil.fasta")

########
#in bash
#cd ../03_build_tree/soil/sequences_soil
#mafft-linsi sequences_soil.fasta > sequences_soil_aligned.fasta #align sequences
#iqtree -s sequences_soil_aligned.fasta #build tree
########
```

```{r soil: show taxa tree, echo=F, message=F, warning=F}

#set to true at the beginning of script if files were computed
if(treefiles){

  #read tree
  tree_file <- "/Users/j.waechli/Desktop/Jan_Waelchli/01_Projects/01_Support/Robin/Microbiome_BE13/03_build_tree/soil/sequences_soil_aligned.fasta.treefile"
  tree <- read.tree(tree_file)
  
  #subset summary and sort them
  summary_taxa <- unique(summary[,c("ASV", "phylum","class","family")])
  summary_taxa <- summary_taxa[summary_taxa$ASV %in% tested_ASVs_soil,]
  rownames(summary_taxa) <- summary_taxa$ASV
  summary_taxa <- summary_taxa[tree[["tip.label"]],]
  
  
  #colorstring
  summary_taxa$labels <- NA
  summary_taxa$cols <- NA
  
  for(a in summary_taxa$ASV){
    summary_taxa$labels[summary_taxa$ASV == a] <- as.character(TAX$labels[TAX$ASV_ID == a][1])
    summary_taxa$cols[summary_taxa$ASV == a] <- as.character(TAX$cols_phyla[TAX$ASV_ID == a][1])
  }
  
  high_abundant_phyla <- as.character(unique(PHYSEQ_phyla_melt$labels_2))
  summary_taxa$labels_2 <- summary_taxa$labels
  summary_taxa$labels_2[!(summary_taxa$labels %in% high_abundant_phyla)] <- "Low abundant phyla"
  summary_taxa$cols2 <- summary_taxa$cols
  summary_taxa$cols2[summary_taxa$labels_2 == "Low abundant phyla"] <- "lightgrey"
  
  #extend taxa string to the length of tree edges (needs for each edge a value but uses only the number of tips)
  phylum <- c(summary_taxa$labels_2,rep(NA,(length(tree[["edge.length"]])-length(summary_taxa$labels_2))+1))
  cols2 <- summary_taxa$cols2
  names(cols2) <- summary_taxa$labels_2
  cols2_long <- c(cols2,rep("white",(length(tree[["edge.length"]])-length(cols2))+1))
  
  #plot genetic tree with phyla
  p1 <- ggtree(tree, layout = "fan", branch.length = "none")+
    geom_tippoint(aes(color=phylum), size=2)+
    scale_color_manual(values=cols2_long)+
    new_scale_color()
  
  
  #for saving
  phylotree_soil <- list(plot=p1, phylum=phylum, colstring=cols2_long)
  
  
  #plot microplastic ring
  df <- summary[summary$Plastic %in% c("CTRL", "microplastic") & summary$ASV %in% rownames(seqs_soil),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_microplastic) / (df$means_CTRL + df$means_microplastic)) *100
  df$abu_trt_microplastic <- factor(df$abu_trt_microplastic, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p2 <- p1 + geom_rect(aes(xmin = 52, xmax = 77, ymin = 1, ymax = nrow(df)), fill="darkolivegreen2")+ #background color
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_microplastic),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list()) +
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  #plot macroplastic ring
  df <- summary[summary$Plastic %in% c("CTRL", "macroplastic") & summary$ASV %in% rownames(seqs_soil),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_macroplastic) / (df$means_CTRL + df$means_macroplastic)) *100
  df$abu_trt_macroplastic <- factor(df$abu_trt_macroplastic, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p3 <- p2 + geom_rect(aes(xmin = 77, xmax = 102, ymin = 1, ymax = nrow(df)), fill="darkseagreen4")+
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_macroplastic),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list())+
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  
  #plot starch ring
  df <- summary[summary$Plastic %in% c("CTRL", "starch") & summary$ASV %in% rownames(seqs_soil),]
  df <- pivot_wider(df,id_cols = !(c(se, score)), names_from = "Plastic", values_from = c("means", "abu_trt", "sensitive"))
  df$difference <- ((df$means_CTRL - df$means_starch) / (df$means_CTRL + df$means_starch)) *100
  df$abu_trt_starch <- factor(df$abu_trt_starch, levels=c("higher in CTRL", "unchanged", "lower in CTRL"))
  
  p4 <- p3 + geom_rect(aes(xmin = 102, xmax = 130, ymin = 1, ymax = nrow(df)), fill="gold")+
              new_scale_fill()+
             geom_fruit(data=df, geom=geom_bar,
              mapping=aes(x=difference, y=ASV, fill=abu_trt_starch),
              pwidth=0.2,
              orientation="y",
              stat="identity",
              offset=0.3,
              axis.params=list(axis="x", text.size=2, text.angle=-90, nbreak=2),
              grid.params=list())+
              scale_fill_manual(values=c("tomato", "grey80", "darkcyan"))
  
  #pseudo plot for legend
  library(ggnewscale)
  df <- data.frame(Phylum=names(cols2), Plastic=c(rep("microplastic",length(cols2)-2),"macroplastic","starch"),
                   Sensitivity=c(rep("higher in CTRL",length(cols2)-2),"unchanged","lower in CTRL"))
  df$Sensitivity <- factor(df$Sensitivity, levels=c("higher in CTRL", "unchanged","lower in CTRL"))
  legend_plot <- ggplot(df,aes(1,1))+
                    geom_bar(stat="identity", aes(fill=Plastic))+
                    geom_point(pch=15, size=6, aes(col=Phylum))+
                    scale_color_manual(values=cols2)+
                    scale_fill_manual(values=level_cols_Plastic[c(1,2,4)], name = "Plastic")+
                    theme_bw()+
                    new_scale_color() +
                    geom_point(pch=15, size=6, aes(col=Sensitivity))+
                    scale_color_manual(values=c("tomato", "grey80", "darkcyan"))
                    #guides(size = guide_legend(order = 1), color = guide_legend(order = 2))
  
  #add plot and legend
  phyla_plot <- plot_grid(p4 + theme(legend.position="none"), 
                          get_legend(legend_plot), nrow=1,rel_widths=c(1, .3))
  phyla_plot
  
}

```

\vspace{5mm}

**Conclusion differential abundance analysis:** *A huge part of the microbiome is shaped by adding plastic to the soil. Microplastic, macroplastic and starch has a similar effect. Bacteria reacting to the plastic are phylogenetic diverse.*


```{r sensitive ASV table, echo=F, message=F, warning=F}

#create supplementary table

#select sensitive ASV and wide transform table
wide_transform <- function(input_data, substrate){
  output <- input_data %>%
  dplyr::mutate(Plastic = paste0(Plastic, "_", substrate)) %>%
  dplyr::mutate(means = round(means,2)) %>% #round
  dplyr::mutate(means = paste0(sprintf("%.2f", means), "%")) %>% #show the same number of digits
  dplyr::mutate(means = ifelse(sensitive == "sensitive", paste0(means," *"), means)) %>%
  dplyr::select(ASV, phylum:genus, means, Plastic) %>%
  pivot_wider(names_from = "Plastic", values_from = "means") %>%
  return(output)
}

summary_sand_wide <- wide_transform(summary_sand, "sand")
summary_root_sand_wide <- wide_transform(summary_root_sand, "root_sand")
summary_root_soil_wide <- wide_transform(summary_root_soil, "root_soil")
summary_soil_wide <- wide_transform(summary_soil, "soil")

#combine the four substrate tables
summary_wide <- dplyr::full_join(summary_sand_wide, summary_root_sand_wide) %>%
  dplyr::full_join(summary_root_soil_wide) %>%
  dplyr::full_join(summary_soil_wide)

# Replace na values with 0%
summary_wide[is.na(summary_wide)] <-  "0.00%"

#filter for sensitive ASV (sensitive to at least one plastic treatment in one substrate)
summary_wide_sensitive <- summary_wide %>%
  dplyr::filter(if_any(CTRL_sand:starch_soil, ~ str_detect(.x, "\\*")))

#xlsx::write.xlsx2(summary_wide_sensitive, "sensitive_ASVs_raw.xlsx")

```


\pagebreak

```{r export RDA files, , echo=F, message=F, warning=F}

# create directory
dir.create("interim")
dir.create("interim/05_DAA")

## set output directory
setwd("interim/05_DAA")

#save objects needed in the following scripts as RDA
# saveRDS(phylotree_sand, "phylotree_sand.RDS")
# saveRDS(phylotree_root_sand, "phylotree_root_sand.RDS")
# saveRDS(phylotree_root_soil, "phylotree_root_soil.RDS")
# saveRDS(phylotree_soil, "phylotree_soil.RDS")

#remove folder created by DAA
unlink("DAA example/", recursive = T)
unlink("05_DAA_files/", recursive = T)

```

